
---
title: "Analysis of Tjornin Data"
author: "Abigail Ramsoe"
output: html_document
---

## Dependencies

Load necessary internal and external libraries 
```{r}
library(DescTools)
library(tidyverse)
library(data.table)
library(taxonomizr)
library(dtplyr)
library(rlang)
library(ggforce)
library(openxlsx)
library(rioja)
library(patchwork)
library(wesanderson)

# internal functions
source("dmg.R")
source("get_calculate_plot_grid.R")
source("perk.R")
source("damage_est_function.R")
source("perk_wrapper.R")
source("perk_wrapper_function.R")
source("get_dmg_decay_fit.R")
source("median.R")
source("filter.R")
```


## Filepaths

We need names, nodes, and mapping of accession to taxid, these don't change unless the reference collection changes 
```{r names-nodes}
# probably don't need to change 
NAMES <- "/projects/caeg/data/db/aeDNA-refs/resources/20230825/ncbi/taxonomy/names.dmp"
NODES <- "/projects/caeg/data/db/aeDNA-refs/resources/20230825/ncbi/taxonomy/nodes.dmp"
ACC2TAXID <- "/projects/lundbeck/scratch/for_antonio/ncbi_taxonomy_01Oct2022/combined_accession2taxid_20221112.gz"
```

Path for result directories (metadmg aggregate and bamfilter stats)
```{r result-dirs}
# change per project
METADMG_DIR <- "/maps/projects/caeg/scratch/iceland/raw_results/stats/metadmg/aggregate/"
BAMFILTER_DIR <- "/maps/projects/caeg/scratch/iceland/raw_results/results/bamfilter/"

# negative data 
NEG_METADMG_DIR <- "/maps/projects/caeg/scratch/iceland/raw_results/stats/metadmg/aggregate/lib/"
NEG_BAMFILTER_DIR <- "/maps/projects/caeg/scratch/iceland/raw_results/results/bamfilter/lib"
```

Path for sequencing metadata
```{r seqdata}
BAM_LIST <- "/maps/projects/caeg/scratch/iceland/library_list.txt"
FC_NAMES <- "/maps/projects/caeg/scratch/iceland/fcnames.csv"
LIB2FC <- "/maps/projects/caeg/scratch/iceland/fc2lib.csv"
```

Metadata tsv file, with sample IDs and dates 
```{r metadata-file}
METADATA <- "/maps/projects/caeg/scratch/iceland/new_metadata.tsv"
```

Places to save output 
```{r output-paths}
PLOT_DIRECTORY <- "/maps/projects/caeg/scratch/iceland/finalplotsV2/"

# have we already run this? if so we can save a lot of time by using save taxonomy info
TAXONOMY_TSV <- "/maps/projects/caeg/scratch/iceland/saved_taxids.tsv" 

# save agg stats here so we can skip this if needed
AGG_TSV <- "/maps/projects/caeg/scratch/iceland/agg_stats.tsv"
NEGAGG_TSV <- "/maps/projects/caeg/scratch/iceland/neg-agg_stats.tsv"

# save damage model here 
GLOBAL_DMG_MODEL <- "/maps/projects/caeg/scratch/iceland/qdata.tsv"

# final excel 
EXCEL_OUT <- "TJORNIN.XLSX" # this is just a basename, it is saved in PLOT_DIRECTORY
```


## Sequencing and Mapping info

Some plots for basic mapping statistics. First we need to load in the bamlist, which shows what libraries we sequenced. 

```{r load-bamlist}
df <- read.csv(BAM_LIST, h = F)
names(df) <- c("bampath")
# split up the bam path to get the CGG number and library code etc 
df$basedir <- sapply(df$bampath, function(x) {
  paste(strsplit(x, "/")[[1]][1:10], collapse = "/")
})
df$label <- sapply(df$bampath, function(x) {
  strsplit(x, "/")[[1]][9]
})
df$lib <- sapply(df$bampath, function(x) {
  strsplit(x, "_")[[1]][5]
})
df$cgg_lib <- paste0(df$label, "_", df$lib)
```

Read in the metadata table, remove the "_" from the CGG number, and save just the "date" and "label" columns, this is all we care about in this case. 
```{r load-metadata}
metadata <- read.table(METADATA,he=T,as.is=T)
metadata$label <- gsub("_", "", metadata$CGG_ID)
names(metadata)[2] <- "date" # we know this column should be called date, this must be changed per project 
metadata <- metadata[, c("label","date")]
metadata$label_fig <- paste0(metadata$label,"-",metadata$date)

head(metadata)
```

Now let's get the sequencing info, which library was sequenced in which run 
```{r load-fcinfo}

fcinfo <- read.csv(LIB2FC,header=F)
names(fcinfo) <- c("fcall","cgg_lib")
fcinfo <- fcinfo %>%
  mutate(
    fc = str_extract(fcall, "^\\d{8}_[A-Z0-9]+")
  )

fcnames <- read.csv(FC_NAMES,header=F)
names(fcnames) <- c("fc", "nicefc")

fcinfo <- inner_join(fcinfo, fcnames)
fcinfo$lib_type <- sapply(fcinfo$nicefc, function(x) {
  strsplit(x, "_")[[1]][3] # SS or DS 
})
fcinfo$label <- sapply(fcinfo$cgg_lib, function(x) {
  strsplit(x, "_")[[1]][1]
})
```

### Plots of number of reads sequenced, trimmed, and assigned

Given a fastqc file, this function gets the total number of reads in the fastq file
```{r fun_totalseqs}
get_total_seqs <- function(file_path) {
  cmd <- paste("unzip -p", shQuote(file_path), "\"*.txt\" | grep \"Total Seq\" | cut -f2")
  total_seqs <- as.numeric(system(cmd, intern = TRUE))
  if (is.na(total_seqs)) {
    return(0)
  }
  return(total_seqs)
}
```

We have multiple lanes per library, each with their own fastq file. So we need to get info for each lane. 
```{r fun_laneseqs}
get_lane_seqs <- function(fastqc_files) {
  lanes <- str_extract(fastqc_files, "L00[1-4]")
  lane_seqs <- sapply(unique(lanes), function(lane) {
    files_by_lane <- fastqc_files[lanes == lane]
    sum(sapply(files_by_lane, get_total_seqs), na.rm = TRUE)
  }, simplify = FALSE)
  return(lane_seqs)
}
```

Now we can use this functionality to get the number of reads in the raw fastq files 
```{r lanedf}
lane_df <- df %>%
  rowwise() %>%
  mutate(
    fastqc_files = list(
      as.character(list.files(
        path = paste0(basedir, "/stats/reads/fastqc_raw/"),
        pattern = "*R1_fastqc.zip$",
        full.names = TRUE
      )) # list of relevant fastqc files per lib 
    ),
    lane_seqs = list(get_lane_seqs(fastqc_files))
  ) %>%
  ungroup()

lane_df <- lane_df %>%
  unnest_longer(lane_seqs) %>% # unnest so we have a new col, lane 
  mutate(lane = names(lane_seqs),  
         total_seqs = unlist(lane_seqs)) %>%  
  select(-lane_seqs, -fastqc_files, -lane_seqs_id) 

head(lane_df)
```

Do the same for the collapsed (post trim) files
```{r colllanedf}
collapsed_lane_df <- df %>%
  rowwise() %>%
  mutate(
    trim_fastqc_files = list(
      as.character(list.files(
        path = paste0(basedir, "/stats/reads/fastqc_trim/"),
        pattern = "*collapsed_fastqc.zip$",
        full.names = TRUE
      ))
    ),
    lane_seqs = list(get_lane_seqs(trim_fastqc_files))
  ) %>%
  ungroup()

collapsed_lane_df <- collapsed_lane_df %>%
  unnest_longer(lane_seqs) %>%  
  mutate(lane = names(lane_seqs),  
         collapsed_seqs = unlist(lane_seqs)) %>%  
  select(-lane_seqs, -trim_fastqc_files, -lane_seqs_id) 
```

Now we can join all the many dataframes we have collected
```{r join}
lane_df <- inner_join(lane_df, collapsed_lane_df)
lane_df <- inner_join(lane_df, fcinfo)
lane_df <- inner_join(lane_df, metadata)
```

Do some plots. Total reads for each cgg number. Each shape is a different lane. We want shapes of the same color (same sequencing, therefore same library) to be very close, otherwise the lanes are imbalanced which would be very strange. 
```{r plot-totreads}
lane_df <- lane_df %>%
  arrange(date) %>%  
  mutate(label = factor(label, levels = unique(label))) # order by date 

ggplot(lane_df, aes(y=label, x=total_seqs, color=nicefc, shape=lane)) +
  geom_point() 
ggsave(paste0(PLOT_DIRECTORY, "/total_reads.png"))
```

Now collapsed reads. 
```{r plot-colreads}
ggplot(lane_df, aes(y=label, x=collapsed_seqs, color=nicefc, shape=lane)) +
  geom_point() 
ggsave(paste0(PLOT_DIRECTORY, "/collapsed_reads.png"))
```

Now that all the lanes seem even, we can sum across the lanes. 
```{r sumlanes}
summed_df <- lane_df %>%
  group_by(label, lib) %>%
  summarize(
    across(c(bampath, basedir, cgg_lib, fcall, date, nicefc), first),
    total_seqs = sum(total_seqs),
    collapsed_seqs = sum(collapsed_seqs),
    .groups = "drop"
  ) %>% 
  arrange(date) %>%  
  mutate(lib = factor(lib, levels = unique(lib)))
```

We can also look at the metadmg results to get the number of assigned (at any taxonomic level) reads
```{r fun_ass}
get_ass_seqs <- function(file_path) {
  cmd <- paste("zcat", shQuote(file_path), "|sed 1d|head -n1|cut -f4")
  total_seqs <- as.numeric(system(cmd, intern = TRUE))
  if (is.na(total_seqs)) {
    return(0)
  }
  return(total_seqs)
}
```

Similar to above, for each library we can find the metadmg result, and get the assigned reads from that 
```{r ass}
ass_df <- df %>%
  rowwise() %>%
  mutate(
    agg_file = 
      as.character(list.files(
        path = paste0(basedir, "/stats/metadmg/aggregate/"),
        pattern = "*_collapsed.stat.gz$",
        full.names = TRUE
      )
    ),
    ass_seqs = get_ass_seqs(agg_file)
  ) %>%
  ungroup()
```

Now we can join that to the previous data about the number of reads summed by lane, and get the percent assigned and collapsed (over the total number of sequenced reads)
```{r getperc}
summed_df <- inner_join(ass_df, summed_df)
summed_df <- summed_df %>%
  mutate(collapsed_pct = (collapsed_seqs / total_seqs) * 100,
  assigned_pct = (ass_seqs / total_seqs) * 100)
```

To plot this we need to convert it to long. Then we can do a nice facet plot showing the number of reads assigned, trimmed and total for each library. 
```{r plot_totalcollass}
long_reads <- summed_df %>% # make looong 
  pivot_longer(
    cols = c(total_seqs, collapsed_seqs, ass_seqs),
    names_to = "read_type",
    values_to = "reads"
  ) %>%
  arrange(date) %>%  
  mutate(lib = factor(lib, levels = unique(lib)))


ggplot(long_reads, aes(y=lib, x=reads, fill=read_type)) +
  geom_bar(stat="identity", position="dodge") + facet_wrap(~nicefc,scales = "free_y")
ggsave(paste0(PLOT_DIRECTORY, "/total_coll_ass.png"))
```

We can do the same, but as boxplots of the percent values 
```{r plot_totalcollass-perc}
long_pct <- summed_df %>%
  pivot_longer(
    cols = c(collapsed_pct, assigned_pct),
    names_to = "read_type_pct",
    values_to = "reads_pct"
  ) %>%
  arrange(date) %>%  
  mutate(label = factor(label, levels = unique(label))) 

ggplot(long_pct, aes(y=label, x=reads_pct, fill=read_type_pct)) +
  geom_boxplot() + facet_wrap(~read_type_pct,scales = "free_x")
ggsave(paste0(PLOT_DIRECTORY, "/perc_coll_ass.png"))
```

### Read length plots of collapsed reads

Now we can do read length distribution plots. First we get the collapsed fastqc files 
```{r a}

collapsed_list <- list()

for (i in 1:nrow(df)) {
  base_dir <- df$basedir[i]
  for (lane in 1:4) { # this might have to change, do we always know this to be the case 
    lane_str <- sprintf("L00%d", lane)    
    collapsed_file <- list.files(path = paste0(base_dir,"/stats/reads/fastqc_trim/"), pattern = paste0(".*", lane_str, "_collapsed_fastqc.zip$"), 
                              full.names = TRUE, recursive = TRUE)    
    collapsed_list <- append(collapsed_list, list(data.frame(
      Lane = lane_str,
      trimmed_collapsed = collapsed_file
    )))
  }
}

collapsed_df <- do.call(rbind, collapsed_list)
df <- cbind(df, collapsed_df)

head(df)
```

Getting the read lengths out of these files is an ordeal. This is how we do it. First get it in a stupid format using this function.
```{r b}
get_read_length <- function(file) {
  cmd <- paste("unzip -p", file, "\"*.txt\" |awk '/#Length/,/>>END_MODULE/'|grep -v L")
  output <- system(cmd, intern = TRUE)
  d <- do.call(rbind, strsplit(output, "\t"))
  d <- as.data.frame(d, stringsAsFactors = FALSE)
  colnames(d) <- c("Lengths", "ReadCount")
  d$ReadCount <- as.numeric(d$ReadCount)
  d$file <- file 
  d$basedir <- sapply(d$file, function(x) {
  paste(strsplit(x, "/")[[1]][1:10], collapse = "/")
  })
  d$label <- sapply(d$file, function(x) {
    strsplit(x, "/")[[1]][9]
  })
  d$lib <- sapply(d$file, function(x) {
    strsplit(x, "_")[[1]][5]
  })
  d$cgg_lib <- paste0(d$label, "_", d$lib)
  d
}
```

And then transform it to a sensible format using this deeply unsensible function 
```{r c}
transform_df <- function(combined_df){
  t <- combined_df %>%
  group_by(cgg_lib, Lengths) %>%
  summarise(TotalReadCount = sum(ReadCount, na.rm = TRUE)) %>%
  ungroup() %>% 
  inner_join(fcinfo, by = "cgg_lib") %>%
  inner_join(metadata) %>% 
  mutate(
    LengthMidpoint = as.numeric(sapply(str_extract_all(Lengths, "\\d+"), function(x) mean(as.numeric(x))))
  ) %>%
  group_by(nicefc) %>%
  mutate(TotalReadsNicefc = sum(TotalReadCount, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(Proportion = TotalReadCount / TotalReadsNicefc) %>%
  select(-Lengths) 
  t
}
```

So like this 
```{r}
all_data <- lapply(df$trimmed_collapsed, get_read_length)
combined_df <- do.call(rbind, all_data)
t <- transform_df(combined_df)
```



Now we can make a beautiful plot
```{r}
ggplot(t, aes(x = LengthMidpoint, y = Proportion, color = date, group = date)) + 
  geom_line() + 
  facet_wrap(~ nicefc) +
  labs(x = "Length (bp)", y = "Proportion", color = "Date") + 
  ggtitle("Collapsed read lengths")
ggsave(paste0(PLOT_DIRECTORY, "/collapsed_readlen.png"))
```

## Read in all data


We need a small helper function to read large files and add the file name itself to a column, we will use this a lot 
```{r fun_read-file}
read_file <- function(f) {
  df <- fread(f, header=T, sep="\t", fill=T, nThread=20)
  df$sample_id <- f
  return(df)
}
```

### Metadmg

Load in metadmg data to a big dataframe. Firstly we get a list of all the metadmg files, and then read them all in to one big dataframe. Lastly, we extract the CGG number from the filepath. 
```{r load-metadmg}
file_list <- list.files(path = METADMG_DIR, pattern = "*gz", recursive = FALSE, full.name=TRUE)
holi_data <- do.call(rbind, lapply(file_list, read_file))
holi_data$label <- sub(".*(CGG[0-9]+)\\..*", "\\1", holi_data$sample_id) # this will have to change when we have non-CGG ids 

head(holi_data)
```



Now we merge the metadmg data with the metadata 
```{r merge-holi-metadata}
holi_data <- inner_join(holi_data, metadata, by = "label")
```

We need a little function to filter the metadmg data. We are only interested in taxa that are classified at species level, and that are Eukaryotes. We make a new column that shows if a taxa is a plant or an animal (based on the taxon path). We then find if the damage profile for each taxa is "good" or "bad" fit to the expected damage model. 
```{r fun_filter-metadmg}
filter_metadmg <- function(df, samples){
  holi_data_sp_euk <- df |>
    filter(rank == "species") |> # must be classified at species level 
    filter(grepl("Eukaryota", taxa_path)) |> # must be a Eukaryote 
    mutate(PlantAnimal = case_when( # define plant/animal based on taxpath 
      grepl("Viridiplantae", taxa_path) ~ "plant",
      grepl("Metazoa", taxa_path) ~ "animal",
    )) |>
    rename(tax_name = taxid, n_reads = nreads) # rename these for ease later

  # get the damage fits using CCC
  dat_all <- dmg_fwd_CCC(holi_data_sp_euk, samples, ci = "asymptotic", nperm = 100, nproc = 24)

  # define good and bad hits
  # bad if confidence interval is above 1 or below 0 (despite other params)
  # good if rho_c > 0.85 and C_b > 0.9 and rho_c p-value < 0.1
  dat_filt <- inner_join(dat_all, holi_data_sp_euk) |>
    mutate(fit = ifelse(rho_c >= 0.85 & C_b > 0.9 & round(rho_c_perm_pval, 3) < 0.1 & !is.na(rho_c), "good", "bad")) |>
    mutate(fit = ifelse(q_CI_h >= 1 | c_CI_l <= 0, "bad", fit))
}
```

Apply this filter on our data 
```{r make-dat_filt}
dat_filt <- suppressWarnings(filter_metadmg(holi_data, metadata$label |> unique())) # we don't want any pesky warnings here 
```


### Bamfilter

Now we are on to the bamfilter data. First we read the names and nodes files 
```{r read-names-nodes}
read.names.sql(NAMES, sqlFile = "nameNode.sqlite", overwrite=TRUE)
read.nodes.sql(NODES, sqlFile = "nameNode.sqlite", overwrite=TRUE)
```

We can read in the data the same way we did for the metadmg results 
```{r load-bamfilter}
file_list <- list.files(path = BAMFILTER_DIR, pattern = "*stats.tsv.gz", recursive = FALSE, full.name=TRUE)
fb_data <- do.call(rbind, lapply(file_list, read_file))
fb_data$label <- sub(".*(CGG[0-9]+)\\..*", "\\1", fb_data$sample_id)

head(fb_data)
```

Now we need to get the taxids that are used in this dataset. If we have already run this analysis, this is saved as a tsv, otherwise we have to find this info from the acc2taxid file 
```{r load-taxids}
if (file.exists(TAXONOMY_TSV)){
	tax_ids <- read_table(TAXONOMY_TSV)
} else { 
	# unique accession IDs 
	accs <- fb_data |>
		select(accession.version = reference) |>
		distinct() |>
		pull(accession.version)

	# read acc2taxid file 
	acc2taxid <- fread(ACC2TAXID, 
		tmpdir = "temp/",
		nThread = 48,
		showProgress = TRUE)

	# filter just those used
	tax_ids <- acc2taxid %>%
		filter(accession.version %in% accs)

	# save for next time 
	write_tsv(tax_ids, TAXONOMY_TSV)
}
```

Then, find the taxonomic info for these taxids 
```{r load-taxoninfo}
# gets the unique taxids
tax_ids_lst <- tax_ids |>
  select(taxid) |>
  distinct() |>
  pull(taxid)

# gets the taxonomic info for the taxids in the data
tax_data <- getTaxonomy(tax_ids_lst, "nameNode.sqlite") |>
  as_tibble() |>
  mutate(taxid = tax_ids_lst)

head(tax_data)
```


## Merge and aggregate to genus level

We need to merge the metadmg and bamfilter data, and also aggregate it to the genus level. Here, we also filter so that taxa must have a median ANI of over 95, an expected breadth ratio of over 0.8, and either gini under 0.6 or entropy above 0.7. Naturally, we remove all taxa with an undefined genus.

```{r aggregate}
agg_stats <- get_stats(tax_ids, tax_data, fb_data, dat_filt, metadata=metadata, mode="library")

if (!file.exists(AGG_TSV)){
	write.csv(agg_stats, AGG_TSV) # write as a csv if not already done 
}
```

## Negative data

We need to pretty much repeat the above exercise with the data from the negative controls. 

Here we read the negative metadmg data
```{r neg-metadmg}
file_list <- list.files(path = NEG_METADMG_DIR, pattern = "*gz", recursive = FALSE, full.name=TRUE)
neg_holi_data <- do.call(rbind, lapply(file_list, read_file))
neg_holi_data$label <- sapply(neg_holi_data$sample_id, function(x) strsplit(x,"_")[[1]][3])
neg_dat_filt <- suppressWarnings(filter_metadmg(neg_holi_data, neg_holi_data$label |> unique()))
```

and the negative bamfilter data 
```{r neg-bamfilter}
file_list <- list.files(path = NEG_BAMFILTER_DIR, pattern = "*stats.tsv.gz", recursive = FALSE, full.name=TRUE)
neg_fb_data <- do.call(rbind, lapply(file_list, read_file))
neg_fb_data$label <- sapply(neg_fb_data$sample_id, function(x) strsplit(x,"_")[[1]][3])
```

and merge and aggregate it to genus level
```{r neg-agg}
neg_agg_stats <- get_stats(tax_ids, tax_data, neg_fb_data, neg_dat_filt, mode="negatives") %>%
  mutate(genus = factor(genus, levels = sort(unique(genus),decreasing=TRUE))) %>% 
  group_by(label) %>%
  ungroup()

if (!file.exists(NEGAGG_TSV)){
	write.csv(neg_agg_stats, NEGAGG_TSV)
}

head(neg_agg_stats)
```

Let's make a nice little heatmap of how many reads are in the negative controls 
```{r plot_neg-heatmap}
ggplot(neg_agg_stats, aes(x = label, y = genus, fill = n_reads)) +
  geom_tile() +
  geom_text(aes(label = n_reads), color = "black", size = 1) +
  scale_fill_gradientn(colors = wes_palette("Zissou1"),values=scales::rescale(c(min(neg_agg_stats$n_reads),median(neg_agg_stats$n_reads),max(neg_agg_stats$n_reads)))) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  ggtitle("Number of reads for each genus in negative controls")
ggsave(paste0(PLOT_DIRECTORY, "/negatives_heatmap.png"))
```

We are interested in the genera with reads in the blanks, so let's add up reads for each library and save this for later
```{r contam-genera}
contam_genera <- neg_agg_stats %>%
	group_by(genus) %>%
	summarise(
	  reads_blanks = sum(n_reads) 
	) %>% arrange(desc(reads_blanks))

head(contam_genera)
```

## Basic plots

### Ranks classified

Let's make some plots. Starting with how many taxa are classified at each taxonomic rank. This only uses metadmg data 

```{r plot_ranks-classified}
# get the top 10 most commonly occuring ranks reads are classified at 
ranks_to_plot <- holi_data %>%
  group_by(rank) %>%
  summarize(totalreads = sum(nreads)) %>%
  arrange(desc(totalreads)) %>% slice_head(n=10) %>% select(rank)

# this is just based on metadmg data, not the aggregated stats 
df <- holi_data %>%
  filter(grepl("Eukaryota", taxa_path)) %>% # get just eukaryotes 
  filter(rank %in% ranks_to_plot$rank) %>% # get ranks we already decided to plot
  group_by(date, rank) %>%
  summarize(numreads = sum(nreads)) %>% # number of reads per rank 
  mutate(proportion = numreads / sum(numreads)) # proportion 

# plot ranks classified (raw numbers)
plot_total <- ggplot(df, aes(x = as.factor(date), y = numreads, fill = rank)) +
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + labs(x = NULL)

# plot proportion
plot_proportion <- ggplot(df, aes(x = as.factor(date), y = proportion, fill = rank)) +
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
  scale_y_continuous(labels = scales::percent)

# combine 
combined_plot <- plot_total / plot_proportion +
    plot_layout(guides = "collect") & theme(legend.position = "bottom")

ggsave(paste0(PLOT_DIRECTORY, "/A.euk.ranks.png"))

print(plot_total)
print(plot_proportion)
```

### Superkingdoms

This also just uses metadmg data. We are interested in the different superkingdoms classified. We can get this info from the taxa path 
```{r}
extract_superkingdom <- function(taxa_path) {
  match <- str_extract(taxa_path, "(?<=:)[^;]+(?=:\"superkingdom\")")
  match <- str_replace_all(match, '[\"/]', '')
  return(ifelse(is.na(match), NA, match))
}

holi_data$superkingdom <- sapply(holi_data$taxa_path, extract_superkingdom)
head(holi_data$superkingdom)
```

Now we can find the total number of reads per superkingdom for each sample. 
```{r}
df_prop <- holi_data %>%
  group_by(label, superkingdom) %>%
  summarise(total_reads = sum(nreads, na.rm = TRUE)) %>%  
  mutate(prop = total_reads / sum(total_reads)) %>% # and also the proportion    
  ungroup() %>% 
  left_join(metadata)
```

Plot it 
```{r}
ggplot(df_prop, aes(y = as.factor(date), x = total_reads, fill = superkingdom)) + 
    geom_bar(stat = "identity") + theme(axis.text=element_text(size=3)) + 
    labs(x = "Number of reads", y = "Library", fill = "Superkingdom", title = " Number of reads classified to each superkingdom")
ggsave(paste0(PLOT_DIRECTORY, "/superkingdom_reads.png"))
```

And plot proportions 
```{r}
ggplot(df_prop, aes(y =  as.factor(date), x = prop, fill = superkingdom)) +
    geom_bar(stat = "identity") + 
    labs(x = "Proportion of reads", y = "Library", fill = "Superkingdom", title =" Proportion of reads classified to each superkingdom")
ggsave(paste0(PLOT_DIRECTORY, "/superkingdom_proportion.png"))
```


### Read length of species level reads

We can look at the mean read lengths per different depth, either for all species level data (good vs bad fit)
```{r}
ggplot(dat_filt, aes(x=mean_rlen, y=as.factor(date))) +
geom_boxplot() +
facet_wrap(~fit)
ggsave(paste0(PLOT_DIRECTORY, "/readlength_species.png"))
```

or for just the plants with >=500 reads 
We can look at the mean read lengths per different depth, either for all species level data (good vs bad fit)
```{r}
ggplot(dat_filt %>% filter(n_reads >= 500 & PlantAnimal == "plant"), aes(x=mean_rlen, y=as.factor(date))) +
geom_boxplot() +
facet_wrap(~fit)
ggsave(paste0(PLOT_DIRECTORY, "/readlength_plants500.png"))

```


## Damage model

Now we start constructing the damage model 

### Good and bad fits

This function makes damage plots for N examples of taxa with at least X reads, seperated into those with good or bad fit to the theoretical damage model.
```{r fun_example-fits}
plot_example_fits <- function(dat, goodbad, plotname="", nreads = 1, howmany){
  tax <- dat |>
    ungroup() |>
    filter(fit == goodbad & n_reads > nreads) |> # get entries with either good or bad fit and over read limit (depending on func arg) 
    group_by(label) |>
    slice_sample(n = howmany) |>
    ungroup()

  samples <- tax$label |> unique()

  plots <- purrr::map(.x = samples, dat = tax, .f = function(x, dat, orient = "fwd", pos = 25, p_breaks = c(0, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6)) {
    data <- dat |> filter(label == x)
    grid_size <- calculate_plot_grid(length(data$tax_name))
    l <- lapply(data$name, function(X) {
      df1 <- data |>
        filter(name == X)
      p <- get_dmg_decay_fit(df1, orient = orient, pos = pos, p_breaks = p_breaks)
      p <- p + ggtitle(X)
      return(p)
    })
    plot <- ggpubr::ggarrange(plotlist = l, ncol = grid_size$cols, nrow = grid_size$rows, align = "hv")
    tit <- paste0(x, " -- ", data$label_fig |> unique())
    ggpubr::annotate_figure(plot, top = ggpubr::text_grob(tit,
                                                          color = "black", face = "bold", size = 12
    ))
  }, .progress = TRUE)

  names(plots) <- samples
  #pdf(plotname, width = 20, height = 20)
  print(plots)
  #dev.off()
}
```

Now we can use that function to make example plots for good and bad fit taxa 
```{r plot_example-fits}
plot_example_fits(dat_filt, "good", paste0(PLOT_DIRECTORY, "/B.good_examples.100reads.pdf"), 100, 10)
plot_example_fits(dat_filt, "bad", paste0(PLOT_DIRECTORY, "/B.bad_examples.100reads.pdf"), 100, 10)
```

### Plot damage in all data

Here we make a plot, for taxa with over 100 reads, of the A_b (dfit/damage) through time. We color the points by good or bad fit, and seperate the plot into two facets - one for plants and one for animals. 
```{r plot_damage-all}
ggplot(agg_stats[agg_stats$n_reads>100,], aes(x = median_A_b, y = date, color = fit, size=n_reads)) +
  geom_point(alpha=0.6) + facet_wrap(~PlantAnimal)
ggsave(paste0(PLOT_DIRECTORY, "/C.dmg.alldata.100reads.png"))
```

### Plot damage in confident plants

Let's make a small function to get the N must abundant genera in plant/animal 
```{r fun-get_top}
get_top <- function(df, n, planimal){
  tmp <- df %>%
  filter(PlantAnimal == planimal) %>%
  group_by(genus) %>%
  summarize(total_reads = sum(n_reads, na.rm = TRUE)) %>%
  arrange(desc(total_reads)) %>%
  slice_head(n = n)
  return(df %>% filter(genus %in% tmp$genus))
}
```

Now we can plot the 20 most abundant plants (only those with at least 500 reads), showing good or bad fit over time 
```{r plot-plants500}
df <- get_top(agg_stats, 20, "plant") %>% filter(n_reads >= 500)
ggplot(df, aes(x = median_A_b, y = date, color = fit, size=n_reads)) +
  geom_point(alpha=0.8) +
  facet_wrap(~genus)
ggsave(paste0(PLOT_DIRECTORY, "/D.dmg.plants.500reads.png"))
```

### Damage model

The actual damage model is coming now. First we need a function that gets the quantiles of the damage pattern over time in the confident plants. For the first 10 and last 10 dates, we allow any taxa to contribute (as we expect less damage at the top of the core, and in the bottom marine layer), but between these, the taxa must have a "good" fit. 
```{r fun-get_condi}
get_conditional_quantile <- function(df) {
	# get 10 first and last dates, these should be filtered differently 
	latest_dates <- df %>%
		distinct(date) %>%
		arrange(date) %>%
		slice(c(1:10, (n()-9):n()))


	# find damage quantiles 
	qdata <- df %>%
		mutate(is_first_dates = date %in% latest_dates$date) %>%
		group_by(date, is_first_dates) %>%
		filter(n() >= 10) %>%
	    summarise(
	      QFILT = if (first(is_first_dates)) {
	        quantile(median_A_b, probs = 0.05, na.rm = TRUE) # keep all taxa 
	      } else {
	        quantile(median_A_b[fit == 'good'], probs = 0.05, na.rm = TRUE) # just good fit for the middle dates 
	      }, .groups = 'drop'
	    ) 

  return(qdata)
}
```

Let's use this to make a line of best fit for the damage over time (based on the damage in plants with over 500 reads). We then get the 5% percentile of this line, to use as a lower limit of the expected damage at each date. 

```{r get-qdata}
# get plants with > 500 reads 
plants500 <- agg_stats %>% filter(PlantAnimal == "plant" & n_reads >= 500)
```
```{r}

# get damage line 
qdata <- get_conditional_quantile(plants500)
loess_fit <- loess(QFILT ~ date, data = qdata, span = 0.3) # loess fit to smooth it 
pred <- predict(loess_fit, newdata = qdata, se = TRUE)
qdata <- qdata %>%
  mutate(
    fit_q = pred$fit,
    lwr = pred$fit - 1.96 * pred$se.fit,
    upr = pred$fit + 1.96 * pred$se.fit,
  ) # get 5% percentile, use this as the limit 
write.table(qdata, GLOBAL_DMG_MODEL)
```

Now we can assign a status (pass or fail) to each taxa. If the median_A_b is below the lower bound of the damage model (5% as calculated earlier), it is fail. If it is above this limit, and has a good fit, it is pass. As previously, for the first 10 and last 10 dates we relax this, so taxa do not need to have a good fit, they simply need to be above the damage lower limit. 

```{r pass-fail}
# assign pass or fail if passed line, if not one of the first or last dates it must also be good fit 
agg_stats <- inner_join(agg_stats, qdata) %>% mutate(status = ifelse(median_A_b < lwr, "fail", "pass"),
  status = ifelse(!is_first_dates & fit == "bad", "fail", status))
```

Now let's plot the damage line and the points that contribute to it 
```{r plot_dmg-model}
# for plotting, assign a low  alpha val to the failed taxa 
plants500 <- agg_stats %>% filter(PlantAnimal == "plant" & n_reads >= 500) %>% 
  mutate(alpha = ifelse(status=="pass",1,0.2))

# plot damage model and the points that made it 
ggplot(plants500) +
  geom_point(aes(y = date, x = median_A_b, color=fit, alpha=alpha)) + # taxa
  geom_ribbon(data = qdata, aes(y = date, xmin = lwr, xmax = upr), alpha = 0.2) + # damage model loess
  geom_path(data = qdata, aes(y = date, x = QFILT)) + # damage model mean 
  geom_path(data = qdata, color="blue", aes(y = date, x = fit_q)) + # damage model real data 
  scale_alpha_identity() +
    labs(y = "Date",
       x = "A_b",
       title = "Damage model")
ggsave(paste0(PLOT_DIRECTORY, "/E.dmgmodel.png"))
```

## Filtering usings the damage model

Now we can use the damage model to filter our data. First we make a little function that helps us filter data. It finds genera with at least N reads at X occurences (and passing the damage filter)
```{r fun_filter-occ}
filter_occ <- function(df, reads, occ){
  filtered_data <- df %>%
  group_by(genus, label) %>%
  filter(n_reads >= reads & status == "pass") %>%
  group_by(genus) %>%
  filter(n_distinct(label) >= occ) %>%
  ungroup()
  unique(filtered_data$genus)
}
```

Find the genera in our data with at least 2 occurences of 100 reads 
```{r good-genera}
good_genera <- filter_occ(agg_stats, 100, 2) # at least 2 occurences of 100 reads, pass
head(good_genera)

agg_stats <- agg_stats %>%
  filter(genus %in% good_genera)
```

### Plot damage patterns in filtered data

Here is a very complicated function to make pages of small damage vs depth plots for each genus, with annotations showing how many each a genus has, and if it has any reads in the blanks 
```{r fun_plot-filtered}
plot_filtered <- function(qdata, df, plotsave, plotname){
  # get annotations about the number of reads in blanks and number of reads in genus
  annotations <- df %>%
    group_by(genus) %>%
    summarise(
      n_reads = sum(n_reads),
      pass_proportion = mean(status == "pass")
    ) %>%
    left_join(contam_genera) %>%
    mutate(reads_blanks = scales::comma(reads_blanks), pass_proportion = scales::percent(pass_proportion), n_reads = scales::comma(n_reads))


  # find out how many pages we need 
  nplots <- length(unique(df$genus))
  rows <- 4
  cols <- 3
  num_pages <- ceiling(nplots / (rows * cols))

  # loop through each page 
  for (page in 1:num_pages) {
    if (nplots < 0) {
      facet_params <- list(facet_wrap_paginate(~genus, page = page)) # if its the last page, we dont care how many rows etc 
    } else {
      facet_params <- list(facet_wrap_paginate(~genus, page = page, nrow = rows, ncol = cols))
    }
  nplots <- nplots - (rows * cols)

  # creating the plot
  plot <- ggplot(df) +
    geom_ribbon(data = qdata, aes(y = date, xmin = lwr, xmax = upr), alpha = 0.4) + # damage model area 
    geom_path(data = qdata, aes(y = date, x = fit_q)) + # damage model mean 
    geom_point(aes(y = date, x = median_A_b, color = status, size = n_reads, alpha = ifelse(status == "pass", 1, 0.4)), show.legend = c(color = TRUE, alpha = FALSE)) +
    scale_alpha_identity() +
    geom_text( 
        data = annotations, # how many reads annotation
            aes(
                label = paste(
                    "Reads: ", n_reads, 
                    "\nPass: ", pass_proportion
                ),
                x = Inf, y = Inf
            ),
            hjust = 1.1, vjust = 1.1,
            inherit.aes = FALSE, size = 3
        ) +
    labs(
      x = "Median Damage (A_b)",
      y = "Date (CE)"
    ) +
     geom_text(
        data = annotations, # reads in blanks annotation 
            aes(
                label = paste(

                    ifelse(!is.na(reads_blanks), paste("\nReads in Blanks: ", reads_blanks), "")
                ),
                x = Inf, y = min(df$date)
            ),
            hjust = 1.1, vjust = 0,
            inherit.aes = FALSE, size = 3, color="red" # red for angry 
        ) +

    xlim(0,0.5)+  
    facet_params[[1]]

  print(plot)

  }
}
```

Let's apply it to plants and animals 
```{r plot-filtered}
# seperate animals and plants 
plant_df <- agg_stats[agg_stats$PlantAnimal == "plant",]
suppressWarnings(plot_filtered(qdata, plant_df, paste0(PLOT_DIRECTORY, "/F.damage.filt.animals.pdf"), "Damage in filtered animals"))

animal_df <- agg_stats[agg_stats$PlantAnimal == "animal",]
suppressWarnings(plot_filtered(qdata, animal_df, paste0(PLOT_DIRECTORY, "/F.damage.filt.plants.pdf"), "Damage in filtered plants"))
```

## Filter for only genera that pass all filters

```{r get-pass}
# get just passing genera 
pass_animal_df <- animal_df[animal_df$status == "pass",]
pass_plant_df <- plant_df[plant_df$status == "pass",]
```

### Plot gini, ani, and breadth

This function does similar page plots, but this time for other variables (given by function arg)
```{r plot_var}
plot_filtered_var <- function(qdata, df, var, plotsave, plotname){

  nplots <- length(unique(df$genus))

  rows <- 4
  cols <- 3
  num_pages <- ceiling(nplots / (rows * cols))


  for (page in 1:num_pages) {
    if (nplots < 0) {
      facet_params <- list(facet_wrap_paginate(~genus, page = page))
    } else {
      facet_params <- list(facet_wrap_paginate(~genus, page = page, nrow = rows, ncol = cols))
    }
    nplots <- nplots - (rows * cols)
    # Creating the plot
    plot <- ggplot(df) +
      geom_point(aes(y = date, x = !!sym(var), color = median_A_b, size = n_reads)) +
      labs(
        y = "Date (CE)"
      ) +
      facet_params[[1]]

    print(plot)

    }
}
```

Now use the function to plot ani, gini and breadth for plants and animals 

#### Animals

##### ANI
```{r var-animal-ani}
plot_filtered_var(qdata, pass_animal_df, "mean_read_ani_median", paste0(PLOT_DIRECTORY, "/G.ani.filt.animals.pdf"), "ANI in filtered animals")
```

##### Breadth
```{r var-animal-breadth}
plot_filtered_var(qdata, pass_animal_df, "penalized_weighted_median_breadth_exp_ratio", paste0(PLOT_DIRECTORY, "/G.breadth.filt.animals.pdf"), "Breadth in filtered animals")
```

##### Gini
```{r var-animal-gini}
plot_filtered_var(qdata, pass_animal_df, "penalized_weighted_median_gini", paste0(PLOT_DIRECTORY, "/G.gini.filt.animals.pdf"), "Gini in filtered animals")
```

#### Plants

##### ANI
```{r var-plant-ani}
plot_filtered_var(qdata, pass_plant_df, "mean_read_ani_median", paste0(PLOT_DIRECTORY, "/G.ani.filt.plants.pdf"), "ANI in filtered plants")
```

##### Breadth
```{r var-plant-breadth}
plot_filtered_var(qdata, pass_plant_df, "penalized_weighted_median_breadth_exp_ratio", paste0(PLOT_DIRECTORY, "/G.breadth.filt.plants.pdf"), "Breadth in filtered plants")
```

##### Gini
```{r var-plant-gini}
plot_filtered_var(qdata, pass_plant_df, "penalized_weighted_median_gini", paste0(PLOT_DIRECTORY, "/G.gini.filt.plants.pdf"), "Gini in filtered plants")
```




## Strat plots

Strat plots show the abundance of genera over time (depth). This little function can make them 
```{r plot_strat}
do_strat_percentage <- function(dat){
  aggregated_data <- dat %>% # get % of each genus at each date (each date sums to 100%)
      group_by(date, genus) %>%
      summarize(n_reads = sum(n_reads, na.rm = TRUE), .groups = 'drop') %>%
      group_by(date) %>%
      mutate(total_reads = sum(n_reads),
             pct_reads = (n_reads / total_reads) * 100) %>%
      ungroup() %>%
      select(date, genus, pct_reads)

  data_wide <- aggregated_data %>% # long to wide 
      pivot_wider(names_from = genus, values_from = pct_reads, values_fill = 0) %>%
      arrange(date)

  data_matrix <- as.matrix(data_wide %>% select(-date)) # we need a matrix 
  rownames(data_matrix) <- data_wide$date

  date_values <- as.numeric(rownames(data_matrix))

  strat.plot(data_matrix, y.rev=FALSE, plot.line=TRUE, plot.poly=FALSE, plot.bar=TRUE,
             lwd.bar=4, sep.bar=TRUE, scale.percent=TRUE, xSpace=0.01,
             x.pc.lab=TRUE, x.pc.omit0=TRUE, srt.xlabel=45, las=2,
             exag=TRUE, exag.mult=5, ylabel = "Date (CE)", yvar = date_values)
}
```

Now make strat plots for animals and plants, with the top 20 and 50 genera 

### Plants

#### 20 top genera
```{r strat-plants20}
do_strat_percentage(get_top(pass_plant_df, 20, "plant"))
```

#### 50 top genera
```{r strat-plants50}
do_strat_percentage(get_top(pass_plant_df, 50, "plant"))
```

### Animals

#### 20 top genera
```{r strat-animals20}
do_strat_percentage(get_top(pass_animal_df, 20, "animal"))
```

#### 50 top genera
```{r strat-animals50}
do_strat_percentage(get_top(pass_animal_df, 50, "animal"))
```



## Excel out

Now we save everything in excel. First we need a function that transforms our dataframes to have columns for each genera, and rows the dates. It will store either percentages (per date), or raw read counts. 
```{r fun-transform}
transform <- function(tibble, calculation = c("sum", "proportion")) {
  calculation <- match.arg(calculation)  

  tibble <- tibble %>%
    select(date, genus, n_reads) %>%  
    group_by(date, genus) %>%        
    summarize(n_reads = sum(n_reads, na.rm = TRUE), .groups = "drop")  

  if (calculation == "proportion") {
    tibble <- tibble %>%
      group_by(date) %>%
      mutate(total_reads = sum(n_reads)) %>%  
      mutate(proportion = n_reads / total_reads) %>%  
      select(-total_reads, -n_reads) %>%  
      pivot_wider(names_from = genus, values_from = proportion)
  } else {
    tibble <- tibble %>%
      pivot_wider(names_from = genus, values_from = n_reads)
  }

  tibble <- tibble %>%
    arrange(desc(date)) %>%             
    select(date, tidyselect::peek_vars()) %>%  
    select(date, order(names(.)[-1]) + 1)  

  return(tibble)
}
```

Transform data, plant and animal with both sum and proportion 
```{r transform}
out_a <- transform(pass_animal_df, "sum")
out_p <- transform(pass_plant_df, "sum")
p_out_a <- transform(pass_animal_df, "proportion")
p_out_p <- transform(pass_plant_df, "proportion")
```

Save as excel
```{r excel}
# create a new workbook
wb <- createWorkbook()

# add worksheets
addWorksheet(wb, "Animals")
writeData(wb, "Animals", out_a)

addWorksheet(wb, "Plants")
writeData(wb, "Plants", out_p)

addWorksheet(wb, "ProportionAnimals")
writeData(wb, "ProportionAnimals", p_out_a)

addWorksheet(wb, "ProportionPlants")
writeData(wb, "ProportionPlants", p_out_p)

addWorksheet(wb, "GeneraInBlanks")
writeData(wb, "GeneraInBlanks", contam_genera) # genera in blanks

# save the workbook to a file
saveWorkbook(wb, paste0(PLOT_DIRECTORY, "/", EXCEL_OUT), overwrite = TRUE)
```

