---
title: "Metagenomic Pipeline Validation"
author: "Mikkel W. Pedersen"
subtitle: "Comparison of test vs. stable pipeline runs"
date: "2025-07-08"
output: html_document
---

# Setup

```{r chunk-1}
library(tidyverse)
library(data.table)
library(dtplyr)
library(rlang)
library(ggforce)
library(viridis)
library(ggh4x)
library(vegan)
library(ggrepel)
library(purrr)
library(eulerr)
library(scales)

# Colorblind-friendly palette (Okabe-Ito)
cb_palette <- c(
  "#E69F00", "#56B4E9", "#009E73", "#F0E442",
  "#0072B2", "#D55E00", "#CC79A7", "#999999"
)

# Global ggplot theme
theme_set(
theme_minimal(base_size = 13) +
  theme(
    plot.title       = element_text(face = "bold", size = 15, margin = margin(b = 6)),
    plot.subtitle    = element_text(color = "grey40", size = 11, margin = margin(b = 10)),
    strip.text       = element_text(face = "bold", size = 11),
    legend.position  = "bottom",
    legend.title     = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    axis.title       = element_text(face = "bold")
  )
)

# Helper: add R-squared annotation to a scatter plot
add_r2 <- function(p, xvar, yvar, data, x_pos = -Inf, y_pos = Inf) {
  valid <- data %>% filter(!is.na({{ xvar }}), !is.na({{ yvar }}))
  r2 <- cor(pull(valid, {{ xvar }}), pull(valid, {{ yvar }}), use = "complete.obs")^2
  p + annotate(
    "text",
    x = x_pos, y = y_pos,
    label = paste0("R2 = ", round(r2, 4)),
    hjust = -0.1, vjust = 1.4,
    size = 4.5, fontface = "bold", color = "grey30"
  )
}

# Human-readable number formatter for axes
label_readable <- function(x) {
  ifelse(
    abs(x) >= 1e6, paste0(round(x / 1e6, 1), "M"),
    ifelse(abs(x) >= 1e3, paste0(round(x / 1e3, 1), "K"),
           format(round(x, 1), big.mark = ","))
  )
}

dir.create("plots", showWarnings = FALSE, recursive = TRUE)
```

```{r chunk-2}
meta <- fread("metadata.csv")
```

```{r chunk-3}
read_file <- function(f) {
  df <- fread(f, header = TRUE, sep = "\t", fill = TRUE, nThread = 2)
  df$sample_id <- f
  return(df)
}

file_list <- list.files(
  path = "/home/jregalado/Projects/GeoGenetics/pipeline_validation/val_unicorn/links",
  pattern = "\\_collapsed.stat.gz$",
  recursive = FALSE,
  full.names = TRUE
)

df <- rbindlist(lapply(file_list, read_file), use.names = TRUE, fill = TRUE)
```

```{r chunk-4}
df <- df %>%
  mutate(
    base        = basename(sample_id),
    comparison  = sapply(strsplit(base, "_"), `[`, 1),
    library_id  = sapply(strsplit(base, "_"), `[`, 3)
  ) %>%
  select(-base)
```

```{r chunk-5}
df <- df %>% left_join(meta)
```

# Root-Level Plots {.tabset}

```{r chunk-6}
root <- df %>% filter(name == "root")
root_wide <- root %>%
  pivot_wider(
    id_cols   = c(names(meta), library_id, taxid, name, rank, lca, taxa_path),
    names_from  = comparison,
    values_from = -c(names(meta), comparison, library_id, taxid, name, rank, lca, taxa_path)
  )
```

## Number of Reads at Root

```{r chunk-7}
p <- ggplot(root_wide, aes(x = nreads_test, y = nreads_stable)) +
  geom_ribbon(
    aes(ymin = nreads_test * 0.9, ymax = nreads_test * 1.1),
    fill = "#56B4E9", alpha = 0.15
  ) +
  geom_abline(color = "grey40", linetype = "dashed", linewidth = 0.6) +
  geom_point(aes(color = label, shape = library_type), size = 3, alpha = 0.85) +
  geom_text_repel(
    data = root_wide %>% filter(abs(nreads_stable - nreads_test) > 0.1 * nreads_test),
    aes(label = library_id),
    size = 3, max.overlaps = 15
  ) +
  scale_color_manual(values = cb_palette) +
  scale_x_continuous(labels = label_readable) +
  scale_y_continuous(labels = label_readable) +
  labs(
    x        = "Number of reads (test pipeline)",
    y        = "Number of reads (stable pipeline)",
    title    = "Number of Reads at Root",
    subtitle = "Shaded band = +/-10% from identity line",
    color    = "Label",
    shape    = "Library type"
  )

p <- add_r2(p, nreads_test, nreads_stable, root_wide)
ggsave("plots/reads_root.pdf", plot = p, width = 10, height = 7)
print(p)
```

## Mean Read Length at Root

```{r chunk-8}
p <- ggplot(root_wide, aes(x = mean_rlen_test, y = mean_rlen_stable)) +
  geom_ribbon(
    aes(ymin = mean_rlen_test * 0.9, ymax = mean_rlen_test * 1.1),
    fill = "#009E73", alpha = 0.15
  ) +
  geom_abline(color = "grey40", linetype = "dashed", linewidth = 0.6) +
  geom_point(aes(color = label, shape = library_type), size = 3, alpha = 0.85) +
  geom_text_repel(
    data = root_wide %>% filter(abs(mean_rlen_stable - mean_rlen_test) > 0.1 * mean_rlen_test),
    aes(label = library_id),
    size = 3, max.overlaps = 15
  ) +
  scale_color_manual(values = cb_palette) +
  labs(
    x        = "Mean read length (test pipeline)",
    y        = "Mean read length (stable pipeline)",
    title    = "Mean Read Length at Root",
    subtitle = "Shaded band = +/-10% from identity line",
    color    = "Label",
    shape    = "Library type"
  )

p <- add_r2(p, mean_rlen_test, mean_rlen_stable, root_wide)
ggsave("plots/rlen_root.pdf", plot = p, width = 10, height = 7)
print(p)
```

## DNA Damage (A) at Root

```{r chunk-9}
p <- ggplot(root_wide, aes(x = A_test, y = A_stable)) +
  geom_ribbon(
    aes(ymin = A_test * 0.9, ymax = A_test * 1.1),
    fill = "#D55E00", alpha = 0.15
  ) +
  geom_abline(color = "grey40", linetype = "dashed", linewidth = 0.6) +
  geom_point(aes(color = label, shape = library_type), size = 3, alpha = 0.85) +
  geom_text_repel(
    data = root_wide %>% filter(abs(A_stable - A_test) > 0.1 * A_test),
    aes(label = library_id),
    size = 3, max.overlaps = 15
  ) +
  scale_color_manual(values = cb_palette) +
  labs(
    x        = "DNA damage A (test pipeline)",
    y        = "DNA damage A (stable pipeline)",
    title    = "DNA Damage (A) at Root",
    subtitle = "Shaded band = +/-10% from identity line",
    color    = "Label",
    shape    = "Library type"
  )

p <- add_r2(p, A_test, A_stable, root_wide)
ggsave("plots/dmg_root.pdf", plot = p, width = 10, height = 7)
print(p)
```

# Superkingdom Distribution {.tabset}

```{r chunk-10}
s <- df %>%
  filter(rank == "phylum") %>%
  mutate(superkingdom = case_when(
    grepl('2759:"Eukaryota":', taxa_path)  ~ "Eukaryota",
    grepl('2:"Bacteria":', taxa_path)      ~ "Bacteria",
    grepl('10239:"Viruses":', taxa_path)   ~ "Viruses",
    grepl('2157:"Archaea":', taxa_path)    ~ "Archaea"
  )) %>%
  group_by(label, comparison, superkingdom) %>%
  summarise(total_nreads = sum(nreads), .groups = "drop")
```

## Raw Read Counts (with differences)

```{r chunk-11}
# Compute differences between test and stable for annotation
s_wide_diff <- s %>%
  pivot_wider(names_from = comparison, values_from = total_nreads, values_fill = 0) %>%
  mutate(
    diff       = test - stable,
    diff_label = paste0(ifelse(diff >= 0, "+", ""), label_readable(diff)),
    y_pos      = pmax(test, stable)
  )

p <- ggplot(s, aes(x = label, y = total_nreads, fill = comparison)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  # Add difference annotations on top
  geom_text(
    data = s_wide_diff,
    aes(x = label, y = y_pos, label = diff_label, fill = NULL),
    vjust = -0.5, size = 3, fontface = "bold", color = "grey30",
    position = position_dodge(width = 0)
  ) +
  facet_wrap(~ superkingdom, scales = "free_y") +
  scale_fill_manual(values = c("stable" = "#0072B2", "test" = "#E69F00")) +
  scale_y_continuous(labels = label_readable, expand = expansion(mult = c(0, 0.15))) +
  labs(
    x     = "Dataset",
    y     = "Total reads",
    fill  = "Pipeline",
    title = "Superkingdom Raw Read Counts",
    subtitle = "Annotations show the difference (test - stable)"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave("plots/superkingdom_nreads.pdf", plot = p, width = 12, height = 8)
print(p)
```

## Normalised Proportions

```{r chunk-12}
df_prop <- s %>%
  group_by(label, comparison) %>%
  mutate(prop_reads = total_nreads / sum(total_nreads)) %>%
  ungroup()

p <- ggplot(df_prop, aes(x = label, y = prop_reads, fill = superkingdom)) +
  geom_col(position = "stack", width = 0.7) +
  facet_wrap(~ comparison) +
  scale_fill_manual(values = c(
    "Eukaryota" = "#009E73",
    "Bacteria"  = "#0072B2",
    "Viruses"   = "#D55E00",
    "Archaea"   = "#CC79A7"
  )) +
  scale_y_continuous(labels = percent_format()) +
  labs(
    x     = "Dataset",
    y     = "Proportion of reads",
    fill  = "Superkingdom",
    title = "Superkingdom Normalised Read Proportions"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave("plots/superkingdom_proportion.pdf", plot = p, width = 10, height = 7)
print(p)
```

# Genus-Level Eukaryotes

```{r chunk-13}
genus <- df %>%
  filter(rank == "genus") %>%
  filter(grepl("Eukaryota", taxa_path)) %>%
  mutate(PlantAnimal = case_when(
    grepl("Viridiplantae", taxa_path) ~ "Plant",
    grepl("Metazoa", taxa_path)       ~ "Animal",
    grepl("Fungi", taxa_path)         ~ "Fungi"
  )) %>%
  filter(!is.na(PlantAnimal))
```

```{r chunk-14}
genus_wide <- genus %>%
  select(-taxa_path) %>%
  pivot_wider(
    names_from  = comparison,
    values_from = -c(names(meta), comparison, library_id, taxid, name, rank, lca, PlantAnimal)
  )
```

## Number of Reads in Eukaryote Genera {.tabset}

### Faceted by Kingdom

Facets are stacked vertically for easier comparison, with readable
log-scale tick labels.

```{r chunk-15}
# Filter to taxa present in both pipelines (nreads > 0 in both) to avoid log10(0)
genus_wide_both <- genus_wide %>%
  filter(!is.na(nreads_test), !is.na(nreads_stable),
         nreads_test > 0, nreads_stable > 0)

n_dropped <- nrow(genus_wide) - nrow(genus_wide_both)
n_only_test   <- sum(!is.na(genus_wide$nreads_test) & genus_wide$nreads_test > 0 &
                     (is.na(genus_wide$nreads_stable) | genus_wide$nreads_stable == 0))
n_only_stable <- sum(!is.na(genus_wide$nreads_stable) & genus_wide$nreads_stable > 0 &
                     (is.na(genus_wide$nreads_test) | genus_wide$nreads_test == 0))

message(paste0(
  "Plotting ", nrow(genus_wide_both), " taxa present in both pipelines. ",
  "Excluded: ", n_dropped, " rows total (",
  n_only_test, " unique to test, ",
  n_only_stable, " unique to stable, rest NA/zero in both)."
))

p <- ggplot(genus_wide_both, aes(x = nreads_test, y = nreads_stable, color = label)) +
  geom_abline(color = "grey40", linetype = "dashed") +
  geom_point(alpha = 0.7, size = 2) +
  scale_y_continuous(trans = "log10", labels = label_readable) +
  scale_x_continuous(trans = "log10", labels = label_readable) +
  scale_color_manual(values = cb_palette) +
  facet_wrap(~ PlantAnimal, scales = "free", ncol = 1) +
  labs(
    x     = "Number of reads (test pipeline)",
    y     = "Number of reads (stable pipeline)",
    title = "Number of Reads in Eukaryote Genera",
    subtitle = paste0("Taxa present in both pipelines; ",
                      n_only_test, " unique to test, ",
                      n_only_stable, " unique to stable (not shown)"),
    color = "Dataset"
  )

ggsave("plots/reads_euk.pdf", plot = p, width = 10, height = 12)
print(p)
```

### Faceted by Kingdom x Label

```{r chunk-16}
p <- ggplot(genus_wide_both, aes(x = nreads_test, y = nreads_stable, color = label)) +
  geom_abline(color = "grey40", linetype = "dashed") +
  geom_point(alpha = 0.7, size = 2) +
  scale_y_continuous(trans = "log10", labels = label_readable) +
  scale_x_continuous(trans = "log10", labels = label_readable) +
  scale_color_manual(values = cb_palette) +
  facet_grid2(PlantAnimal ~ label, scales = "free") +
  labs(
    x     = "Number of reads (test pipeline)",
    y     = "Number of reads (stable pipeline)",
    title = "Number of Reads in Eukaryote Genera (detailed)",
    color = "Dataset"
  )

ggsave("plots/reads_euk_facet.pdf", plot = p, width = 14, height = 12)
print(p)
```

## Volcano Plot

```{r chunk-17}
summary_df <- genus %>%
  filter(comparison %in% c("test", "stable")) %>%
  group_by(label, PlantAnimal, taxid, name, comparison) %>%
  summarise(total_nreads = sum(nreads), .groups = "drop")

wide_df <- summary_df %>%
  filter(total_nreads >= 10) %>%
  pivot_wider(names_from = comparison, values_from = total_nreads)

wide_df <- wide_df %>%
  mutate(log2FC = log2((test + 1) / (stable + 1)))

pvals <- genus %>%
  filter(comparison %in% c("test", "stable")) %>%
  group_by(label, PlantAnimal, taxid, name) %>%
  summarise(pval = tryCatch(
    wilcox.test(nreads ~ comparison)$p.value,
    error = function(e) NA_real_
  ), .groups = "drop")

volcano_df <- wide_df %>%
  left_join(pvals, by = c("label", "PlantAnimal", "taxid", "name")) %>%
  mutate(log10p = -log10(pval))

volcano_df$avg_nreads <- rowMeans(
  cbind(
    ifelse(is.na(volcano_df$test), 0, volcano_df$test),
    ifelse(is.na(volcano_df$stable), 0, volcano_df$stable)
  )
)

p <- ggplot(volcano_df, aes(x = log2FC, y = log10p, color = avg_nreads, size = avg_nreads)) +
  geom_point(alpha = 0.7) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "grey50") +
  geom_hline(yintercept = -log10(0.05), linetype = "dotted", color = "red") +
  scale_colour_viridis_c(trans = "log10", option = "C", labels = label_readable) +
  scale_size_continuous(range = c(1, 5), guide = "none") +
  facet_grid2(label ~ PlantAnimal, scales = "free") +
  labs(
    x     = "log2 Fold Change (test vs stable)",
    y     = "-log10(p-value)",
    title = "Volcano Plot of Read Counts",
    subtitle = "Only taxa with >=10 reads; dashed lines = 2-fold change; dotted red = p = 0.05",
    color = "Avg. reads"
  )

ggsave("plots/reads_euk_volcano.pdf", plot = p, width = 12, height = 12)
print(p)
```

### Top Differentially Represented Taxa

```{r chunk-18}
top_hits <- volcano_df %>%
  filter(!is.na(log2FC), !is.na(pval)) %>%
  filter(pval < 0.05) %>%
  group_by(label, PlantAnimal) %>%
  slice_max(order_by = abs(log2FC), n = 10, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(name = fct_reorder(name, log2FC))

for (lab in unique(top_hits$label)) {
  p <- ggplot(top_hits %>% filter(label == lab), aes(x = name, y = log2FC, fill = PlantAnimal)) +
    geom_col(width = 0.7) +
    facet_wrap(~ PlantAnimal, scales = "free") +
    coord_flip() +
    scale_fill_manual(values = c("Plant" = "#009E73", "Animal" = "#0072B2", "Fungi" = "#CC79A7")) +
    labs(
      title = paste0("Top Differentially Represented Taxa - ", lab),
      x     = "Taxon",
      y     = "log2 Fold Change (test vs stable)",
      fill  = "Kingdom"
    )

  ggsave(paste0("plots/reads_euk_volcano_diff_", lab, ".pdf"), plot = p, width = 10, height = 7)
  print(p)
  cat("\n\n")
}
```

## Taxa Gain/Loss

The original code took slice_head(n = 30) per
PlantAnimal group, so with 3 groups you could get up to 90 bars. The fix
below takes the top 30 across all groups per label.

```{r chunk-19}
read_diff <- genus %>%
  select(name, taxid, library_id, comparison, nreads, PlantAnimal, label) %>%
  pivot_wider(names_from = comparison, values_from = nreads) %>%
  mutate(read_change = test - stable)

for (l in unique(df$label)) {

  d <- read_diff %>%
    filter(label == l) %>%
    filter(!is.na(read_change)) %>%
    # Take top 30 overall (not per group!) by absolute change
    arrange(desc(abs(read_change))) %>%
    slice_head(n = 30) %>%
    mutate(direction = ifelse(read_change >= 0, "Gain in test", "Loss in test"))

  p <- ggplot(d, aes(x = reorder(name, read_change), y = read_change, fill = direction)) +
    geom_col(width = 0.7) +
    coord_flip() +
    scale_fill_manual(values = c("Gain in test" = "#009E73", "Loss in test" = "#D55E00")) +
    scale_y_continuous(labels = label_readable) +
    labs(
      x        = "Taxon",
      y        = "Read count difference (test - stable)",
      fill     = "Direction",
      title    = paste0("Taxa Gain/Loss - ", l),
      subtitle = "Top 30 taxa by absolute read count change"
    )

  ggsave(paste0("plots/taxa_gainloss_", l, ".pdf"), plot = p, width = 10, height = 9)
  print(p)
  cat("\n\n")
}
```

## Damage Patterns {.tabset}

### Damage Comparison PDF

This generates a large multi-page PDF of all damage pattern
comparisons (taxa with >=100 reads in both pipelines).

```{r chunk-20}
plot_pipeline_comparison <- function(df2, only5 = FALSE) {

  nreads_old <- df2 %>% filter(comparison == "stable") %>% pull(nreads)
  nreads_new <- df2 %>% filter(comparison == "test")   %>% pull(nreads)
  A_b_old    <- df2 %>% filter(comparison == "stable") %>% pull(A)
  A_b_new    <- df2 %>% filter(comparison == "test")   %>% pull(A)
  lib_id     <- df2$library_id[1]
  lab        <- df2$label[1]
  taxon      <- df2$name[1]
  tid        <- df2$taxid[1]

  x_cols <- grep("^fwdx\\d+$", names(df2), value = TRUE)
  f_cols <- grep("^fwf\\d+$",  names(df2), value = TRUE)

  long_df <- bind_rows(
    df2 %>% filter(comparison == "stable") %>%
      select(all_of(x_cols), all_of(f_cols)) %>%
      pivot_longer(everything(), names_to = "metric", values_to = "value") %>%
      mutate(position = as.integer(str_extract(metric, "\\d+")),
             type = ifelse(str_starts(metric, "fwdx"), "x", "f"),
             comparison = "stable"),
    df2 %>% filter(comparison == "test") %>%
      select(all_of(x_cols), all_of(f_cols)) %>%
      pivot_longer(everything(), names_to = "metric", values_to = "value") %>%
      mutate(position = as.integer(str_extract(metric, "\\d+")),
             type = ifelse(str_starts(metric, "fwdx"), "x", "f"),
             comparison = "test")
  ) %>%
    filter(!is.na(value)) %>%
    mutate(type = factor(type, levels = c("f", "x")))

  if (only5) long_df <- long_df %>% filter(position < 7)

  A_points <- tibble(
    position   = 0,
    value      = c(A_b_old, A_b_new),
    type       = "A",
    comparison = c("stable", "test")
  )

  title <- paste0(
    taxon, " (taxid: ", tid, ")\n",
    "Library: ", lib_id, ", Label: ", lab, "\n",
    "nreads (stable/test): ", nreads_old, " / ", nreads_new,
    ", A: ", round(A_b_old, 3), " / ", round(A_b_new, 3)
  )

  ggplot(long_df, aes(x = position, y = value, color = comparison, linetype = type)) +
    geom_line() +
    geom_point(data = filter(long_df, type == "f")) +
    geom_point(data = A_points, aes(shape = type), size = 3, stroke = 1.2, inherit.aes = TRUE) +
    scale_color_manual(values = c("stable" = "#0072B2", "test" = "#E69F00")) +
    scale_linetype_manual(values = c("x" = "solid", "f" = "dashed")) +
    scale_shape_manual(values = c(A = 8)) +
    labs(
      title    = title,
      x        = "Position from 5' end",
      y        = "Damage Estimate",
      color    = "Pipeline",
      linetype = "Metric",
      shape    = NULL
    )
}
```

```{r chunk-21}
# Which taxa are only in one pipeline version?
taxa_by_pipeline <- genus %>%
  group_by(name, taxid, library_id, label) %>%
  summarise(
    in_stable = any(comparison == "stable"),
    in_test   = any(comparison == "test"),
    .groups   = "drop"
  )

only_stable <- taxa_by_pipeline %>% filter(in_stable & !in_test)
only_test   <- taxa_by_pipeline %>% filter(!in_stable & in_test)

message(paste0(
  "Taxa in both pipelines: ", sum(taxa_by_pipeline$in_stable & taxa_by_pipeline$in_test), "\n",
  "Taxa only in stable: ", nrow(only_stable), "\n",
  "Taxa only in test: ", nrow(only_test)
))
```

#### Taxa unique to stable pipeline

```{r chunk-22}
if (nrow(only_stable) > 0) {
  only_stable_detail <- genus %>%
    semi_join(only_stable, by = c("name", "taxid", "library_id")) %>%
    filter(comparison == "stable") %>%
    select(library_id, label, name, taxid, nreads, A) %>%
    arrange(desc(nreads))
  knitr::kable(only_stable_detail, caption = paste0("Taxa found only in stable (n=", nrow(only_stable_detail), ")"))
} else {
  message("No taxa unique to stable pipeline.")
}
```

#### Taxa unique to test pipeline

```{r chunk-23}
if (nrow(only_test) > 0) {
  only_test_detail <- genus %>%
    semi_join(only_test, by = c("name", "taxid", "library_id")) %>%
    filter(comparison == "test") %>%
    select(library_id, label, name, taxid, nreads, A) %>%
    arrange(desc(nreads))
  knitr::kable(only_test_detail, caption = paste0("Taxa found only in test (n=", nrow(only_test_detail), ")"))
} else {
  message("No taxa unique to test pipeline.")
}
```

```{r chunk-24}
valid_taxa <- genus %>%
  group_by(name, taxid, library_id) %>%
  filter(all(c("stable", "test") %in% comparison)) %>%
  summarise(
    n_old = nreads[comparison == "stable"][1],
    n_new = nreads[comparison == "test"][1],
    .groups = "drop"
  ) %>%
  filter(n_old >= 100, n_new >= 100)

genus_filtered <- semi_join(genus, valid_taxa, by = c("name", "taxid", "library_id"))

groups <- genus_filtered %>%
  group_by(name, taxid, library_id) %>%
  group_split()

plots <- list()
for (df2 in groups) {
  p <- plot_pipeline_comparison(df2)
  plots <- append(plots, list(p))
}

if (length(plots) > 0) {
  pdf("damage_patterns.pdf", width = 8, height = 6)
  for (p in plots) print(p)
  dev.off()
  message(paste(length(plots), "damage pattern plots saved to damage_patterns.pdf"))
} else {
  message("No valid plots generated - PDF not created.")
}
```

### DNA Damage (A) in Eukaryotes

```{r chunk-25}
dmg_data <- genus_wide %>% filter(nreads_test >= 100 & nreads_stable >= 100)

p <- ggplot(dmg_data, aes(x = A_test, y = A_stable, color = label)) +
  geom_abline(color = "grey40", linetype = "dashed") +
  geom_point(alpha = 0.7, size = 2.5) +
  scale_color_manual(values = cb_palette) +
  facet_wrap(~ PlantAnimal, ncol = 1) +
  labs(
    x     = "DNA damage A (test)",
    y     = "DNA damage A (stable)",
    title = "DNA Damage (A) in Eukaryote Genera",
    subtitle = "Only genera with >=100 reads in both pipelines",
    color = "Dataset"
  )

# Per-facet R2 annotation
for (grp in unique(dmg_data$PlantAnimal)) {
  sub <- dmg_data %>% filter(PlantAnimal == grp, !is.na(A_test), !is.na(A_stable))
  if (nrow(sub) > 2) {
    r2 <- round(cor(sub$A_test, sub$A_stable, use = "complete.obs")^2, 4)
    p <- p + annotate(
      "text",
      x = min(dmg_data$A_test, na.rm = TRUE),
      y = max(dmg_data$A_stable, na.rm = TRUE),
      label = paste0("R2 = ", r2),
      hjust = -0.1, vjust = 1.4,
      size = 4, fontface = "bold", color = "grey30"
    )
  }
}

ggsave("plots/dmg_euk.pdf", plot = p, width = 10, height = 10)
print(p)
```

## Total Number of Reads per Kingdom

Facets stacked vertically with R2 per facet and readable axis
numbers.

```{r chunk-26}
id_df <- genus %>%
  group_by(library_id, label, PlantAnimal, comparison) %>%
  summarise(total_nreads = sum(nreads), .groups = "drop") %>%
  pivot_wider(names_from = comparison, values_from = total_nreads)

# Pre-compute R2 per kingdom for annotation
r2_labels <- id_df %>%
  group_by(PlantAnimal) %>%
  summarise(
    r2   = round(cor(test, stable, use = "complete.obs")^2, 4),
    x    = min(test, na.rm = TRUE),
    y    = max(stable, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(label_text = paste0("R2 = ", r2))

p <- ggplot(id_df, aes(x = test, y = stable, color = label)) +
  geom_abline(color = "grey40", linetype = "dashed") +
  geom_point(size = 3, alpha = 0.85) +
  geom_text(
    data = r2_labels,
    aes(x = x, y = y, label = label_text),
    inherit.aes = FALSE,
    hjust = -0.1, vjust = 1.4,
    size = 4.5, fontface = "bold", color = "grey30"
  ) +
  scale_color_manual(values = cb_palette) +
  scale_x_continuous(labels = label_readable) +
  scale_y_continuous(labels = label_readable) +
  facet_wrap(~ PlantAnimal, scales = "free", ncol = 1) +
  labs(
    x     = "Total reads (test pipeline)",
    y     = "Total reads (stable pipeline)",
    title = "Total Number of Reads per Kingdom",
    color = "Dataset"
  )

ggsave("plots/reads_total_euk.pdf", plot = p, width = 10, height = 12)
print(p)
```

## Total Number of Taxa per Kingdom

```{r chunk-27}
taxa_df <- genus %>%
  group_by(library_id, label, PlantAnimal, comparison) %>%
  summarise(num = n(), .groups = "drop") %>%
  pivot_wider(names_from = comparison, values_from = num)

r2_taxa <- taxa_df %>%
  group_by(PlantAnimal) %>%
  summarise(
    r2 = round(cor(test, stable, use = "complete.obs")^2, 4),
    x  = min(test, na.rm = TRUE),
    y  = max(stable, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(label_text = paste0("R2 = ", r2))

p <- ggplot(taxa_df, aes(x = test, y = stable, color = label)) +
  geom_abline(color = "grey40", linetype = "dashed") +
  geom_point(size = 3, alpha = 0.85) +
  geom_text(
    data = r2_taxa,
    aes(x = x, y = y, label = label_text),
    inherit.aes = FALSE,
    hjust = -0.1, vjust = 1.4,
    size = 4.5, fontface = "bold", color = "grey30"
  ) +
  scale_color_manual(values = cb_palette) +
  facet_wrap(~ PlantAnimal, scales = "free", ncol = 1) +
  labs(
    x     = "Number of taxa (test pipeline)",
    y     = "Number of taxa (stable pipeline)",
    title = "Total Number of Taxa per Kingdom",
    color = "Dataset"
  )

ggsave("plots/taxa_euk.pdf", plot = p, width = 10, height = 12)
print(p)
```

# Unique Taxa Overlap

Summary of taxa overlap between stable and test pipelines per library
(>=100 reads).

```{r chunk-28}
library_ids <- unique(genus$library_id)

overlap_summary <- list()

for (lib in library_ids) {
  df_lib <- genus %>% filter(library_id == lib & nreads >= 100)
  lab <- unique(df_lib$label)[1]

  for (group in unique(df_lib$PlantAnimal)) {
    df_group <- df_lib %>% filter(PlantAnimal == group)

    taxa_stable <- df_group %>% filter(comparison == "stable") %>% pull(name) %>% unique()
    taxa_test   <- df_group %>% filter(comparison == "test")   %>% pull(name) %>% unique()

    n_both       <- length(intersect(taxa_stable, taxa_test))
    n_only_stable <- length(setdiff(taxa_stable, taxa_test))
    n_only_test   <- length(setdiff(taxa_test, taxa_stable))

    overlap_summary <- append(overlap_summary, list(tibble(
      library_id   = lib,
      label        = lab,
      kingdom      = group,
      shared       = n_both,
      only_stable  = n_only_stable,
      only_test    = n_only_test,
      total_stable = length(taxa_stable),
      total_test   = length(taxa_test)
    )))
  }
}

overlap_df <- bind_rows(overlap_summary)
```

```{r chunk-29}
knitr::kable(
  overlap_df,
  caption = "Taxa overlap per library and kingdom (>=100 reads)",
  col.names = c("Library", "Dataset", "Kingdom", "Shared", "Only stable",
                "Only test", "Total stable", "Total test")
)
```

```{r chunk-30}
overlap_long <- overlap_df %>%
  select(library_id, label, kingdom, shared, only_stable, only_test) %>%
  pivot_longer(
    cols      = c(shared, only_stable, only_test),
    names_to  = "category",
    values_to = "count"
  ) %>%
  mutate(
    category = factor(category,
      levels = c("only_stable", "shared", "only_test"),
      labels = c("Only in stable", "Shared", "Only in test")
    ),
    lib_label = paste0(library_id, " (", label, ")")
  )

p <- ggplot(overlap_long, aes(x = lib_label, y = count, fill = category)) +
  geom_col(position = "stack", width = 0.7) +
  coord_flip() +
  facet_wrap(~ kingdom, scales = "free_x", ncol = 1) +
  scale_fill_manual(values = c(
    "Only in stable" = "#0072B2",
    "Shared"         = "#999999",
    "Only in test"   = "#E69F00"
  )) +
  labs(
    x     = "Library",
    y     = "Number of taxa (>=100 reads)",
    fill  = "Presence",
    title = "Taxa Overlap Between Pipelines",
    subtitle = "Per library and kingdom, taxa with >=100 reads"
  )

ggsave("plots/taxa_overlap.pdf", plot = p, width = 12, height = max(8, length(library_ids) * 0.5))
print(p)
```

```{r chunk-31}
# Save Euler/Venn diagrams to a separate PDF (these don't render inline reliably)
library_ids <- unique(genus$library_id)
venn_plots_all <- list()

for (lib in library_ids) {
  df_lib <- genus %>% filter(library_id == lib & nreads >= 100)
  lab <- unique(df_lib$label)[1]

  venn_plots <- list()
  for (group in unique(df_lib$PlantAnimal)) {
    df_group <- df_lib %>% filter(PlantAnimal == group)

    taxa_stable <- df_group %>% filter(comparison == "stable") %>% pull(name) %>% unique()
    taxa_test   <- df_group %>% filter(comparison == "test")   %>% pull(name) %>% unique()

    if (length(taxa_stable) == 0 & length(taxa_test) == 0) next

    venn_counts <- list(stable = taxa_stable, test = taxa_test)
    fit <- tryCatch(euler(venn_counts), error = function(e) NULL)

    if (!is.null(fit)) {
      p <- plot(
        fit,
        fills = list(fill = c("#0072B2", "#E69F00"), alpha = 0.5),
        quantities = TRUE,
        main = paste(lib, "|", group)
      )
      venn_plots <- append(venn_plots, list(p))
    }
  }

  if (length(venn_plots) > 0) {
    combined <- do.call(
      gridExtra::arrangeGrob,
      c(venn_plots, ncol = min(3, length(venn_plots)),
        top = paste("Taxa overlap (>=100 reads) -", lib, "-", lab))
    )
    venn_plots_all <- append(venn_plots_all, list(combined))
  }
}

if (length(venn_plots_all) > 0) {
  pdf("venn_taxa_by_library.pdf", width = 12, height = 8)
  for (p in venn_plots_all) {
    grid::grid.draw(p)
    grid::grid.newpage()
  }
  dev.off()
  message(paste(length(venn_plots_all), "Venn diagram pages saved to venn_taxa_by_library.pdf"))
} else {
  message("No Venn diagrams generated.")
}
```

# PCoA {.tabset}

## Regular PCoA

```{r chunk-32}
plots <- list()

for (lab in unique(genus$label)) {
  for (group in c("Plant", "Animal", "Fungi")) {

    genus_sub <- genus %>%
      filter(label == lab, PlantAnimal == group, !is.na(name), !is.na(nreads)) %>%
      mutate(
        taxa   = paste0(name, "-", taxid),
        sample = paste(library_id, comparison, sep = "_")
      )

    if (length(unique(genus_sub$sample)) < 3) next

    tmp_dt     <- as.data.table(genus_sub)
    wide_counts <- dcast(tmp_dt, taxa ~ sample, value.var = "nreads", fill = 0)

    if (nrow(wide_counts) < 2) next

    rownames(wide_counts) <- wide_counts$taxa
    wide_data <- as.data.frame(wide_counts[, -1])

    dist_matrix <- vegdist(t(wide_data), method = "bray")
    pcoa_result <- cmdscale(dist_matrix, k = 2, eig = TRUE)

    pcoa_df <- data.frame(pcoa_result$points)
    colnames(pcoa_df) <- c("PCoA1", "PCoA2")
    pcoa_df$sample <- rownames(pcoa_result$points)

    pcoa_df <- pcoa_df %>%
      separate(sample, into = c("library_id", "comparison"), sep = "_", remove = FALSE) %>%
      left_join(
        genus_sub %>% select(library_id, comparison) %>% distinct(),
        by = c("library_id", "comparison")
      )

    p <- ggplot(pcoa_df, aes(x = PCoA1, y = PCoA2, color = comparison)) +
      geom_path(aes(group = library_id), color = "grey60", linewidth = 0.5, alpha = 0.7) +
      geom_point(size = 4) +
      scale_color_manual(values = c("stable" = "#0072B2", "test" = "#E69F00")) +
      labs(
        title = paste("PCoA -", lab, "-", group),
        x     = "PCoA 1",
        y     = "PCoA 2",
        color = "Pipeline"
      )

    plots <- append(plots, list(p))
  }
}

if (length(plots) > 0) {
  pdf("pcoa.pdf", width = 10, height = 8)
  for (p in plots) print(p)
  dev.off()

  for (p in plots) { print(p); cat("\n\n") }
}
```

## PCoA Influence (Leave-One-Out)

Which taxa drive the biggest difference between pipeline runs?

```{r chunk-33}
plots <- list()

for (lab in unique(genus$label)) {
  for (group in c("Plant", "Animal", "Fungi")) {

    df2 <- genus %>%
      filter(label == lab, PlantAnimal == group, !is.na(name), !is.na(nreads)) %>%
      mutate(
        taxa   = paste0(name, "-", taxid),
        sample = paste(library_id, comparison, sep = "_")
      )

    if (length(unique(df2$sample)) < 3) next

    tmp_dt     <- as.data.table(df2)
    wide_counts <- dcast(tmp_dt, taxa ~ sample, value.var = "nreads", fill = 0)

    if (nrow(wide_counts) < 2) next

    rownames(wide_counts) <- wide_counts$taxa
    wide_data <- as.data.frame(wide_counts[, -1])
    rownames(wide_data) <- wide_counts$taxa

    dist_matrix <- vegdist(t(wide_data), method = "bray")
    pcoa       <- cmdscale(dist_matrix, k = 2)
    coord_full <- data.frame(pcoa)
    coord_full$sample <- rownames(coord_full)

    coord_full <- coord_full %>%
      separate(sample, into = c("library_id", "comparison"), sep = "_", remove = FALSE)

    baseline_diffs <- coord_full %>%
      pivot_wider(id_cols = library_id, names_from = comparison, values_from = c(X1, X2)) %>%
      mutate(euclidean = sqrt((X1_stable - X1_test)^2 + (X2_stable - X2_test)^2)) %>%
      select(library_id, baseline = euclidean)

    taxa_list <- rownames(wide_data)
    results   <- list()

    for (tax in taxa_list) {
      reduced_data <- wide_data[setdiff(rownames(wide_data), tax), ]
      dist_reduced <- vegdist(t(reduced_data), method = "bray")
      pcoa_reduced <- cmdscale(dist_reduced, k = 2)

      coords <- data.frame(pcoa_reduced)
      coords$sample <- rownames(coords)
      coords <- coords %>%
        separate(sample, into = c("library_id", "comparison"), sep = "_", remove = FALSE)

      diffs <- coords %>%
        pivot_wider(id_cols = library_id, names_from = comparison, values_from = c(X1, X2)) %>%
        mutate(euclidean = sqrt((X1_stable - X1_test)^2 + (X2_stable - X2_test)^2)) %>%
        select(library_id, euclidean) %>%
        left_join(baseline_diffs, by = "library_id") %>%
        mutate(taxa_removed = tax, delta = baseline - euclidean)

      results[[tax]] <- diffs
    }

    influence_df <- bind_rows(results)

    summary_df <- influence_df %>%
      group_by(taxa_removed) %>%
      summarise(mean_delta = mean(delta, na.rm = TRUE), .groups = "drop") %>%
      arrange(desc(mean_delta))

    filtered_df <- summary_df %>%
      arrange(mean_delta) %>%
      slice(c(1:10, (n() - 9):n()))

    p <- ggplot(filtered_df, aes(
      x    = reorder(taxa_removed, mean_delta),
      y    = mean_delta,
      fill = mean_delta > 0
    )) +
      geom_col(width = 0.7) +
      coord_flip() +
      scale_fill_manual(
        values = c("TRUE" = "#009E73", "FALSE" = "#D55E00"),
        labels = c("TRUE" = "Reduces diff.", "FALSE" = "Increases diff."),
        name   = "Effect"
      ) +
      labs(
        title = paste("PCoA Influencers -", lab, "-", group),
        x     = "Taxon removed",
        y     = "Mean Delta PCoA distance"
      )

    plots <- append(plots, list(p))
  }
}

if (length(plots) > 0) {
  pdf("pcoa_influence.pdf", width = 10, height = 8)
  for (p in plots) print(p)
  dev.off()

  for (p in plots) { print(p); cat("\n\n") }
}
```

# Taxa Shared Across Datasets {.tabset}

## Ubiquitous Taxa (present in all datasets)

```{r chunk-34}
all_labels     <- unique(genus$label)
n_total_labels <- length(all_labels)

abundant_taxa <- genus %>% filter(nreads >= 100)

ubiquitous_taxa <- abundant_taxa %>%
  group_by(name, PlantAnimal, comparison) %>%
  summarise(
    n_labels    = n_distinct(label),
    total_reads = sum(nreads),
    .groups     = "drop"
  ) %>%
  filter(n_labels == n_total_labels)

p <- ggplot(ubiquitous_taxa, aes(x = reorder(name, total_reads), y = total_reads, fill = PlantAnimal)) +
  geom_col(width = 0.7) +
  coord_flip() +
  scale_fill_manual(values = c("Plant" = "#009E73", "Animal" = "#0072B2", "Fungi" = "#CC79A7")) +
  scale_y_continuous(trans = "log10", labels = label_readable) +
  facet_wrap(~ comparison) +
  labs(
    title = "Taxa Found in All Datasets (>=100 reads)",
    x     = "Taxon (genus)",
    y     = "Total reads (all datasets combined)",
    fill  = "Kingdom"
  )

ggsave("plots/ubiquitous_taxa.pdf", plot = p, width = 12, height = 8)
print(p)
```

## Damage in Ubiquitous Taxa

```{r chunk-35}
bad <- genus %>% filter(name %in% ubiquitous_taxa$name)

# Use distinct shapes + colors for better differentiation
p <- ggplot(bad, aes(x = nreads, y = A, shape = comparison, color = name)) +
  geom_point(size = 3, alpha = 0.8) +
  scale_x_continuous(trans = "log10", labels = label_readable) +
  scale_color_manual(values = rep(cb_palette, length.out = length(unique(bad$name)))) +
  scale_shape_manual(values = c("stable" = 16, "test" = 17)) +
  labs(
    title = "Damage in Ubiquitous Taxa (>=100 reads)",
    x     = "Number of reads",
    y     = "DNA damage (A)",
    color = "Taxon",
    shape = "Pipeline"
  )

ggsave("plots/ubiquitous_taxa_dmg.pdf", plot = p, width = 12, height = 7)
print(p)
```

## Top 20 Taxa Appearing Across Multiple Datasets

Improved with colorblind-friendly fill scales and clearer bar
layouts.

```{r chunk-36}
taxa_presence <- abundant_taxa %>%
  group_by(name, PlantAnimal, comparison) %>%
  summarise(
    n_labels    = n_distinct(label),
    total_reads = sum(nreads),
    .groups     = "drop"
  )

taxa_filtered <- taxa_presence %>% filter(n_labels >= 2)

for (pl in unique(taxa_filtered$PlantAnimal)) {
  for (s in c("stable", "test")) {

    df_sub <- taxa_filtered %>% filter(PlantAnimal == pl & comparison == s)

    if (pl == "Plant") {
      df_sub <- df_sub %>% filter(total_reads >= 500 & n_labels >= 3)
    }

    if (nrow(df_sub) == 0) next

    # Take top 20 for readability
    df_sub <- df_sub %>%
      arrange(desc(n_labels), desc(total_reads)) %>%
      slice_head(n = 20)

    p <- ggplot(df_sub, aes(
      x    = reorder(name, n_labels),
      y    = n_labels,
      fill = log10(total_reads + 1)
    )) +
      geom_col(width = 0.7) +
      coord_flip() +
      scale_fill_viridis_c(option = "D", name = "log10(reads)") +
      labs(
        title = paste0(pl, " taxa across datasets - ", s, " pipeline"),
        x     = "Taxon (genus)",
        y     = "Number of datasets"
      )

    fname <- paste0("plots/common_taxa_", pl, "_", s, ".pdf")
    ggsave(fname, plot = p, width = 10, height = 8)
    print(p)
    cat("\n\n")
  }
}
```

# Ratio Heatmaps

Ratio heatmap comparing test/stable read counts per genus across
datasets, to catch systematic biases between pipelines.

```{r chunk-37}
# Build a matrix of log2(test/stable) ratios per taxon per label
ratio_data <- genus %>%
  select(name, taxid, library_id, label, comparison, nreads, PlantAnimal) %>%
  pivot_wider(names_from = comparison, values_from = nreads) %>%
  filter(!is.na(test), !is.na(stable)) %>%
  filter(test >= 10 | stable >= 10) %>%
  mutate(log2_ratio = log2((test + 1) / (stable + 1)))

for (grp in unique(ratio_data$PlantAnimal)) {

  d <- ratio_data %>%
    filter(PlantAnimal == grp) %>%
    group_by(name) %>%
    mutate(mean_abs_ratio = mean(abs(log2_ratio))) %>%
    ungroup()

  # Keep top 30 most variable taxa
  top_taxa <- d %>%
    distinct(name, mean_abs_ratio) %>%
    arrange(desc(mean_abs_ratio)) %>%
    slice_head(n = 30) %>%
    pull(name)

  d_top <- d %>% filter(name %in% top_taxa)

  if (nrow(d_top) == 0) next

  p <- ggplot(d_top, aes(
    x    = paste(library_id, label, sep = "\n"),
    y    = reorder(name, mean_abs_ratio),
    fill = log2_ratio
  )) +
    geom_tile(color = "white", linewidth = 0.3) +
    scale_fill_gradient2(
      low      = "#0072B2",
      mid      = "grey95",
      high     = "#D55E00",
      midpoint = 0,
      name     = "log2(test/stable)"
    ) +
    labs(
      title = paste("Read Count Ratio Heatmap -", grp),
      subtitle = "Top 30 most variable taxa across libraries",
      x     = "Library",
      y     = "Taxon"
    ) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
      axis.text.y = element_text(size = 9)
    )

  ggsave(paste0("plots/ratio_heatmap_", grp, ".pdf"), plot = p, width = 14, height = 10)
  print(p)
  cat("\n\n")
}
```

# Zfit and sigmaD Comparison

The test pipeline produces Zfit_old,
Zfit_new, sigmaD_old, and
sigmaD_new, while the stable pipeline produces
Zfit and sigmaD. Here we compare these values
for matching taxa (by library_id and
name).

```{r chunk-38, eval=TRUE}
# Extract stable values: Zfit and sigmaD
stable_vals <- df %>%
  filter(comparison == "stable") %>%
  select(library_id, name, taxid, label, PlantAnimal = rank,
         Zfit_stable = Zfit, sigmaD_stable = sigmaD)

# Extract test values: Zfit_old, Zfit_new, sigmaD_old, sigmaD_new
test_vals <- df %>%
  filter(comparison == "test") %>%
  select(library_id, name, taxid, label,
         Zfit_old, Zfit_new, sigmaD_old, sigmaD_new)

# Join on library_id + name (taxon)
compare_df <- inner_join(stable_vals, test_vals, by = c("library_id", "name", "taxid", "label"))

message(paste0(
  nrow(compare_df), " taxa matched between stable and test pipelines across ",
  n_distinct(compare_df$library_id), " libraries."
))
```

## Zfit: stable vs test (old and new) {.tabset}

### Zfit stable vs Zfit_old (test)

```{r chunk-39, eval=TRUE}
d <- compare_df %>% filter(!is.na(Zfit_stable), !is.na(Zfit_old))

p <- ggplot(d, aes(x = Zfit_old, y = Zfit_stable)) +
  geom_abline(color = "grey40", linetype = "dashed", linewidth = 0.6) +
  geom_point(aes(color = label), alpha = 0.5, size = 1.5) +
  scale_color_manual(values = cb_palette) +
  labs(
    x     = "Zfit_old (test pipeline)",
    y     = "Zfit (stable pipeline)",
    title = "Zfit: stable vs Zfit_old from test",
    color = "Dataset"
  )

p <- add_r2(p, Zfit_old, Zfit_stable, d)
ggsave("plots/zfit_stable_vs_old.pdf", plot = p, width = 10, height = 7)
print(p)
```

### Zfit stable vs Zfit_new (test)

```{r chunk-40, eval=TRUE}
d <- compare_df %>% filter(!is.na(Zfit_stable), !is.na(Zfit_new))

p <- ggplot(d, aes(x = Zfit_new, y = Zfit_stable)) +
  geom_abline(color = "grey40", linetype = "dashed", linewidth = 0.6) +
  geom_point(aes(color = label), alpha = 0.5, size = 1.5) +
  scale_color_manual(values = cb_palette) +
  labs(
    x     = "Zfit_new (test pipeline)",
    y     = "Zfit (stable pipeline)",
    title = "Zfit: stable vs Zfit_new from test",
    color = "Dataset"
  )

p <- add_r2(p, Zfit_new, Zfit_stable, d)
ggsave("plots/zfit_stable_vs_new.pdf", plot = p, width = 10, height = 7)
print(p)
```

### Zfit_old vs Zfit_new (within test)

```{r chunk-41, eval=TRUE}
d <- compare_df %>% filter(!is.na(Zfit_old), !is.na(Zfit_new))

p <- ggplot(d, aes(x = Zfit_old, y = Zfit_new)) +
  geom_abline(color = "grey40", linetype = "dashed", linewidth = 0.6) +
  geom_point(aes(color = label), alpha = 0.5, size = 1.5) +
  scale_color_manual(values = cb_palette) +
  labs(
    x     = "Zfit_old (test pipeline)",
    y     = "Zfit_new (test pipeline)",
    title = "Zfit_old vs Zfit_new (within test pipeline)",
    color = "Dataset"
  )

p <- add_r2(p, Zfit_old, Zfit_new, d)
ggsave("plots/zfit_old_vs_new.pdf", plot = p, width = 10, height = 7)
print(p)
```

## sigmaD: stable vs test (old and new) {.tabset}

### sigmaD stable vs sigmaD_old (test)

```{r chunk-42, eval=TRUE}
d <- compare_df %>% filter(!is.na(sigmaD_stable), !is.na(sigmaD_old))

p <- ggplot(d, aes(x = sigmaD_old, y = sigmaD_stable)) +
  geom_abline(color = "grey40", linetype = "dashed", linewidth = 0.6) +
  geom_point(aes(color = label), alpha = 0.5, size = 1.5) +
  scale_color_manual(values = cb_palette) +
  labs(
    x     = "sigmaD_old (test pipeline)",
    y     = "sigmaD (stable pipeline)",
    title = "sigmaD: stable vs sigmaD_old from test",
    color = "Dataset"
  )

p <- add_r2(p, sigmaD_old, sigmaD_stable, d)
ggsave("plots/sigmad_stable_vs_old.pdf", plot = p, width = 10, height = 7)
print(p)
```

### sigmaD stable vs sigmaD_new (test)

```{r chunk-43, eval=TRUE}
d <- compare_df %>% filter(!is.na(sigmaD_stable), !is.na(sigmaD_new))

p <- ggplot(d, aes(x = sigmaD_new, y = sigmaD_stable)) +
  geom_abline(color = "grey40", linetype = "dashed", linewidth = 0.6) +
  geom_point(aes(color = label), alpha = 0.5, size = 1.5) +
  scale_color_manual(values = cb_palette) +
  labs(
    x     = "sigmaD_new (test pipeline)",
    y     = "sigmaD (stable pipeline)",
    title = "sigmaD: stable vs sigmaD_new from test",
    color = "Dataset"
  )

p <- add_r2(p, sigmaD_new, sigmaD_stable, d)
ggsave("plots/sigmad_stable_vs_new.pdf", plot = p, width = 10, height = 7)
print(p)
```

### sigmaD_old vs sigmaD_new (within test)

```{r chunk-44, eval=TRUE}
d <- compare_df %>% filter(!is.na(sigmaD_old), !is.na(sigmaD_new))

p <- ggplot(d, aes(x = sigmaD_old, y = sigmaD_new)) +
  geom_abline(color = "grey40", linetype = "dashed", linewidth = 0.6) +
  geom_point(aes(color = label), alpha = 0.5, size = 1.5) +
  scale_color_manual(values = cb_palette) +
  labs(
    x     = "sigmaD_old (test pipeline)",
    y     = "sigmaD_new (test pipeline)",
    title = "sigmaD_old vs sigmaD_new (within test pipeline)",
    color = "Dataset"
  )

p <- add_r2(p, sigmaD_old, sigmaD_new, d)
ggsave("plots/sigmad_old_vs_new.pdf", plot = p, width = 10, height = 7)
print(p)
```

## Distribution of differences

```{r chunk-45, eval=TRUE}
diff_df <- compare_df %>%
  mutate(
    dZfit_old  = Zfit_old - Zfit_stable,
    dZfit_new  = Zfit_new - Zfit_stable,
    dsigmaD_old  = sigmaD_old - sigmaD_stable,
    dsigmaD_new  = sigmaD_new - sigmaD_stable
  ) %>%
  select(library_id, name, label, starts_with("dZfit"), starts_with("dsigmaD")) %>%
  pivot_longer(
    cols      = -c(library_id, name, label),
    names_to  = "metric",
    values_to = "diff"
  ) %>%
  filter(!is.na(diff)) %>%
  mutate(
    param   = ifelse(grepl("Zfit", metric), "Zfit", "sigmaD"),
    version = ifelse(grepl("_old", metric), "old - stable", "new - stable")
  )

p <- ggplot(diff_df, aes(x = diff, fill = version)) +
  geom_density(alpha = 0.5) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  scale_fill_manual(values = c("old - stable" = "#0072B2", "new - stable" = "#E69F00")) +
  facet_wrap(~ param, scales = "free", ncol = 1) +
  labs(
    x     = "Difference (test - stable)",
    y     = "Density",
    title = "Distribution of Zfit and sigmaD differences",
    subtitle = "Centered on zero means perfect agreement between pipelines",
    fill  = "Comparison"
  )

ggsave("plots/zfit_sigmad_diff_density.pdf", plot = p, width = 10, height = 10)
print(p)
```

# DUST Score Diagnostics (test pipeline only)

DUST measures read-set complexity (0-100, where <7 is high
complexity and higher values indicate low complexity / repetitive
sequence). These columns are only present in the test pipeline.

```{r chunk-46, eval=TRUE}
# Filter to test pipeline rows that have DUST values
dust_df <- df %>%
  filter(comparison == "test", !is.na(mean_dust)) %>%
  left_join(
    genus %>% select(library_id, name, taxid, PlantAnimal) %>% distinct(),
    by = c("library_id", "name", "taxid")
  )

message(paste0(nrow(dust_df), " rows with DUST scores in test pipeline"))
```

## DUST vs Mean GC Content

```{r chunk-47, eval=TRUE}
p <- ggplot(dust_df, aes(x = mean_gc, y = mean_dust, color = label, size = nreads)) +
  geom_point(alpha = 0.4) +
  geom_hline(yintercept = 7, linetype = "dashed", color = "red") +
  scale_size_continuous(range = c(0.5, 4), trans = "log10", labels = label_readable, name = "nreads") +
  scale_color_manual(values = cb_palette) +
  annotate("text", x = max(dust_df$mean_gc, na.rm = TRUE), y = 7,
           label = "DUST = 7 (high complexity threshold)",
           hjust = 1, vjust = -0.5, size = 3.5, color = "red") +
  labs(
    x     = "Mean GC content",
    y     = "Mean DUST score",
    title = "DUST Score vs GC Content",
    subtitle = "Lower DUST = higher complexity; dashed line = complexity threshold (7)",
    color = "Dataset"
  )

ggsave("plots/dust_vs_gc.pdf", plot = p, width = 10, height = 7)
print(p)
```

## DUST vs Mean Read Length

```{r chunk-48, eval=TRUE}
p <- ggplot(dust_df, aes(x = mean_rlen, y = mean_dust, color = label, size = nreads)) +
  geom_point(alpha = 0.4) +
  geom_hline(yintercept = 7, linetype = "dashed", color = "red") +
  scale_size_continuous(range = c(0.5, 4), trans = "log10", labels = label_readable, name = "nreads") +
  scale_color_manual(values = cb_palette) +
  labs(
    x     = "Mean read length (bp)",
    y     = "Mean DUST score",
    title = "DUST Score vs Read Length",
    subtitle = "Lower DUST = higher complexity; dashed line = complexity threshold (7)",
    color = "Dataset"
  )

ggsave("plots/dust_vs_rlen.pdf", plot = p, width = 10, height = 7)
print(p)
```

## DUST vs Number of Reads

```{r chunk-49, eval=TRUE}
p <- ggplot(dust_df, aes(x = nreads, y = mean_dust, color = label)) +
  geom_point(alpha = 0.4, size = 1.5) +
  geom_hline(yintercept = 7, linetype = "dashed", color = "red") +
  scale_x_continuous(trans = "log10", labels = label_readable) +
  scale_color_manual(values = cb_palette) +
  labs(
    x     = "Number of reads",
    y     = "Mean DUST score",
    title = "DUST Score vs Number of Reads",
    subtitle = "Lower DUST = higher complexity; dashed line = complexity threshold (7)",
    color = "Dataset"
  )

ggsave("plots/dust_vs_nreads.pdf", plot = p, width = 10, height = 7)
print(p)
```

## DUST Multi-Panel Overview

All three relationships in one figure, faceted by rank for the
eukaryote genera.

```{r chunk-50, eval=TRUE}
dust_euk <- dust_df %>% filter(!is.na(PlantAnimal))

if (nrow(dust_euk) > 0) {
  dust_long <- dust_euk %>%
    select(name, library_id, label, PlantAnimal, mean_dust, mean_gc, mean_rlen, nreads) %>%
    pivot_longer(
      cols      = c(mean_gc, mean_rlen, nreads),
      names_to  = "metric",
      values_to = "value"
    ) %>%
    mutate(metric = case_when(
      metric == "mean_gc"   ~ "GC content",
      metric == "mean_rlen" ~ "Read length (bp)",
      metric == "nreads"    ~ "Number of reads"
    ))

  p <- ggplot(dust_long, aes(x = value, y = mean_dust, color = label)) +
    geom_point(alpha = 0.3, size = 1.2) +
    geom_hline(yintercept = 7, linetype = "dashed", color = "red") +
    scale_color_manual(values = cb_palette) +
    facet_grid(PlantAnimal ~ metric, scales = "free_x") +
    labs(
      x     = "",
      y     = "Mean DUST score",
      title = "DUST Score by Kingdom and Metric (eukaryotes only)",
      color = "Dataset"
    )

  ggsave("plots/dust_multipanel.pdf", plot = p, width = 14, height = 12)
  print(p)
}
```

# Species-per-Read Diagnostics (test pipeline only)

Mean number of species per read (mean_nspec) indicates taxonomic
ambiguity: higher values mean reads map to more species, suggesting less
confident assignments.

```{r chunk-51, eval=TRUE}
nspec_df <- df %>%
  filter(comparison == "test", !is.na(mean_nspec))

message(paste0(nrow(nspec_df), " rows with nspec values in test pipeline"))
```

## Mean Species-per-Read vs Number of Alignments

```{r chunk-52, eval=TRUE}
p <- ggplot(nspec_df, aes(x = nalign, y = mean_nspec, color = label, size = nreads)) +
  geom_point(alpha = 0.4) +
  scale_x_continuous(trans = "log10", labels = label_readable) +
  scale_size_continuous(range = c(0.5, 4), trans = "log10", labels = label_readable, name = "nreads") +
  scale_color_manual(values = cb_palette) +
  labs(
    x     = "Number of alignments (nalign)",
    y     = "Mean species per read",
    title = "Taxonomic Ambiguity vs Number of Alignments",
    subtitle = "Higher mean_nspec = more species per read = less specific assignment",
    color = "Dataset"
  )

ggsave("plots/nspec_vs_nalign.pdf", plot = p, width = 10, height = 7)
print(p)
```

## Mean Species-per-Read vs Number of Reads

```{r chunk-53, eval=TRUE}
p <- ggplot(nspec_df, aes(x = nreads, y = mean_nspec, color = label)) +
  geom_point(alpha = 0.4, size = 1.5) +
  scale_x_continuous(trans = "log10", labels = label_readable) +
  scale_color_manual(values = cb_palette) +
  labs(
    x     = "Number of reads",
    y     = "Mean species per read",
    title = "Taxonomic Ambiguity vs Read Depth",
    color = "Dataset"
  )

ggsave("plots/nspec_vs_nreads.pdf", plot = p, width = 10, height = 7)
print(p)
```

## Species-per-Read Distribution by Dataset

```{r chunk-54, eval=TRUE}
p <- ggplot(nspec_df, aes(x = mean_nspec, fill = label)) +
  geom_density(alpha = 0.4) +
  scale_fill_manual(values = cb_palette) +
  labs(
    x     = "Mean species per read",
    y     = "Density",
    title = "Distribution of Mean Species-per-Read Across Taxa",
    subtitle = "Higher values indicate more taxonomic ambiguity",
    fill  = "Dataset"
  )

ggsave("plots/nspec_density.pdf", plot = p, width = 10, height = 7)
print(p)
```

## Variance in Species-per-Read vs Mean

```{r chunk-55, eval=TRUE}
p <- ggplot(nspec_df, aes(x = mean_nspec, y = var_nspec, color = label)) +
  geom_point(alpha = 0.4, size = 1.5) +
  scale_color_manual(values = cb_palette) +
  labs(
    x     = "Mean species per read",
    y     = "Variance in species per read",
    title = "Mean vs Variance of Species-per-Read",
    subtitle = "High variance + high mean = inconsistent, ambiguous assignments",
    color = "Dataset"
  )

ggsave("plots/nspec_mean_vs_var.pdf", plot = p, width = 10, height = 7)
print(p)
```

```{r chunk-56}
sessionInfo()
```
