---
title: "Validation"
output: html_document
date: "2025-07-08"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Intro 

We want to compare a new "version" of the pipeline (this could be a new reference catalog, new mapping software, or new parameters etc) with the current "stable version". For this we run the validation pipeline with the validation libraries (metadata [here](https://docs.google.com/spreadsheets/d/16m6KTU2PfVh_40NzRZMKmTlEiArrnWHE4dMbYiGF4n8/edit?gid=0#gid=0)).


# Setup

This relies on some bash magic that we used server side to rename the files before download, so that the name is e.g. `stable_Lib_LV123_collapsed.stat.gz`, and symlinks the results to a folder which can then be downloaded using rsync. We will have to edit the MD5 sum identifiers for test and stable each time here, depending on what we want to compare. This example shows comparing the v1 reference catalog (stable) with the v3 catalog (test).

```{bash eval=FALSE}
#ssh dandycomp01fl
for line in $(find /projects/caeg/data/validation -name '*_collapsed.stat.gz' | grep 'v1.0.3' | grep 'results/metadmg/aggregate')
do   
  status=""
  if echo "$line" | grep -q "eebd93d8414a08f40337174bb00ee3dc"; then # if this is in path, that means it's the new catalog
      status="test"
  fi
  if echo "$line" | grep -q "48ca4558f97def8149639ed5c1b40c37"; then # this is v1
      status="stable"
  fi
  if [ ! -z $status ]; then 
    ln -s "$line" "/home/dlm551/validation_links/${status}_$(basename $line)" 
  fi 
done 
```

Now we have a server-side folder of symlinks to the aggregated stats, we can download it with rsync 

```{bash eval=FALSE}
rsync -av dlm551@dandycomp01fl:/home/dlm551/validation_links/ . -L
```

We also need to download the metadata. Download from [here](https://docs.google.com/spreadsheets/d/16m6KTU2PfVh_40NzRZMKmTlEiArrnWHE4dMbYiGF4n8/edit?gid=0#gid=0) and save as `metadata.csv`.

# Read in data to R

Start with libraries
```{r, message=FALSE}
library(tidyverse)
library(data.table)
library(dtplyr)
library(rlang)
library(ggforce)
library(wesanderson)
library(viridis)
library(ggh4x)
library(vegan)
library(ggrepel)
library(purrr)
library(eulerr)
```

Get the metadata, remember to download this from the google doc linked above
```{r}
meta <- fread("metadata.csv")
```

We need all the aggregated stat files that we prepared and downloaded earlier
```{r}
# small function to read file and add the file name as a column
read_file <- function(f) {
  df <- fread(f, header=T, sep="\t", fill=T, nThread=2)
  df$sample_id <- f
  return(df)
}

# list all files that were downloaded
file_list <- list.files(
  path = "aggregate/v1_v3", # this will need to be changed, depending on directory structure
  pattern = "\\.gz$",  
  recursive = FALSE,
  full.names = TRUE
)

# read the files into a big df 
df <- do.call(rbind, lapply(file_list, read_file))
```

We can use the file name in the df to get the comparison type (stable or test), and the library id 
```{r}
df <- df %>%
  mutate(
    base = basename(sample_id),
    comparison = sapply(strsplit(base, "_"), `[`, 1),
    library_id  = sapply(strsplit(base, "_"), `[`, 3)
  ) %>%
  select(-base)
```

Join this with the metadata
```{r}
df <- df %>% left_join(meta, by = "library_id")
```

The final df here is a long df of the combined aggregated stats for all libraries and all comparisons (both test and stable), the comparison column distinguishes these from each other. 


# Root classifications

We start with just statistics for reads classified at root level. Since we are working with the aggregated stats, this is summed up the tree, so this also includes reads classified anywhere from subspecies to root. 

```{r}
root <- df %>% filter(name == "root") # just root 

# make wide for easy plotting
root_wide <- root %>%
  pivot_wider( 
    id_cols = c(names(meta), c(library_id, taxid, name, rank, lca)),
    names_from = comparison,
    values_from = -c(names(meta), c(comparison,library_id, taxid, name, rank, lca, taxa_path))
  )
```


## Root - Number of assigned reads 

First, we plot the number of reads assigned total.
```{r, warning=FALSE}
p <- ggplot(root_wide, aes(x = nreads_test, y = nreads_stable)) +
  geom_abline(color = "blue", linetype = "dashed") +
  geom_point(aes(color=label, shape=library_type),size = 5) +
  labs(
    x = "number of reads (log) - test",
    y = "number of reads (log) - stable",
    title = "Number of reads at root",
    shape = "library type",
    color = "dataset"
  ) + 
  scale_x_continuous(trans="log10") +
  scale_y_continuous(trans="log10") + 
  theme_test(base_size = 15) 
ggsave("plots/reads_root.pdf", plot=p)

print(p)
```

If points fall below the diagonal line, the library gets more reads assigned in the test pipeline. If points fall above, the library loses reads in the test pipeline. Look to see if there are systematic difference (e.g. all libraries from a certain dataset, or a certain library type, are different between the two runs).

## Root - Average read length 

Now we plot the average length of the assigned reads at root
```{r, warning=FALSE}
p <- ggplot(root_wide, aes(x = mean_rlen_test, y = mean_rlen_stable)) +
  geom_abline(color = "blue", linetype = "dashed") +
  geom_point(aes(color=label, shape=library_type), size = 5) +
  labs(
    x = "mean read length - test",
    y = "mean read length - stable",
    title = "Mean read length at root",
    shape = "library type",
    color = "dataset"
  ) + 
    theme_test(base_size = 15) 
ggsave("plots/rlen_root.pdf", plot=p)
print(p)
```

If points fall below the diagonal line, the library has a longer average read length in the test pipeline. If points fall above, the reads are shorter in the test pipeline. 

## Root - Average DNA damage


```{r, warning=FALSE}
p <- ggplot(root_wide, aes(x = A_test, y = A_stable)) +
  geom_abline(color = "blue", linetype = "dashed") +
  geom_point(aes(color=label, shape=library_type), size = 5) +
  labs(
    x = "A - test",
    y = "A - stable",
    title = "DNA damage (A) at root",
    shape = "library type",
    color = "dataset"
  ) + 
    theme_test(base_size = 15) 

ggsave("plots/dmg_root.pdf", plot=p)
print(p)
```
If points fall below the diagonal line, the library has higher average damage in the test pipeline. If points fall above, the damage is less in the test pipeline. 


# Superkingdom distribution

## Superkingdom raw counts 

Butterfly pipelines look at eukaryotes, but reads at root may contain any superkingdom. Here we plot the distribution of these. Remember some pipelines have microbial pre-filtering, where we remove bacterial reads.
```{r}
# assign superkingdom by an ugly grep on the taxpath
# here we end up with just read count per superkingdom, per comparison (stable, test), and per dataset
s <- df %>% filter(rank == "phylum") %>% # get this at a high level 
    mutate(superkingdom = case_when( 
      grepl('2759:"Eukaryota":', taxa_path) ~ "eukaryota",
      grepl('2:"Bacteria":', taxa_path) ~ "bacteria",
      grepl('10239:"Viruses":', taxa_path) ~ "virus",
      grepl('2157:"Archaea":', taxa_path) ~ "archaea")) %>% 
  group_by(label, comparison, superkingdom) %>% 
  summarise(total_nreads = sum(nreads)) 

# plot with facets of superkingdom, so we compare stable and test 
p <- ggplot(s, aes(x = label, y = total_nreads, fill = comparison)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8)) +
  facet_wrap(~superkingdom, scales="free") +
  labs(
    x = "dataset",
    y = "total reads",
    fill = "version",
    title = "Superkingdom raw read counts"
  ) +
  theme_test(base_size = 15) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 

ggsave("plots/superkingdom_nreads.pdf",plot=p)
print(p)

```

If the bars are different sizes between stable and test for the same dataset, it means that a given superkingdom is either more or less abundant in the test pipeline. As mentioned above, this could also show different pre-treatments (e.g. microbial filtering), or different reference catalogs (e.g. including a microbe reference)


## Superkingdom normalised counts 

We are also interested in the relative abundance of each superkingdom
```{r}
# get the proportion out of the number of total reads 
df_prop <- s %>%
  group_by(label, comparison) %>%
  mutate(prop_reads = total_nreads / sum(total_nreads)) %>%
  ungroup()

p <- ggplot(df_prop, aes(x = label, y = prop_reads, fill = superkingdom)) +
  geom_bar(stat = "identity", position = "stack") +
  facet_wrap(~comparison) +
  labs(
    x = "dataset",
    y = "proportion of reads",
    fill = "superkingdom",
    title = "Superkingdom normalised counts"
  ) +
  scale_y_continuous(labels = scales::percent_format()) +
  theme_test(base_size = 15) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 

ggsave("plots/superkingdom_proportion.pdf",plot=p)
print(p)
```

Here we see the overall proportion of superkingdoms in each dataset, across the two comparison statuses. 


# Genus level eukaryotes 

Let's say we only care about eukaryotes (because it's true). Let's just focus on those identified at a genus level 

First prepare a dataframe full of the genus level eukaryotes, and make the column PlantAnimal (plant, animal, fungi). 
```{r}
genus <- df %>% filter(rank == "genus") |> # must be classified at genus level 
    filter(grepl("Eukaryota", taxa_path)) |> # must be a Eukaryote 
    mutate(PlantAnimal = case_when( # define plant/animal based on taxpath 
      grepl("Viridiplantae", taxa_path) ~ "plant",
      grepl("Metazoa", taxa_path) ~ "animal",
      grepl("Fungi", taxa_path) ~ "fungi")) %>% 
  filter(!is.na(PlantAnimal))

# make a wide version too
# here we remove taxa path, as there can be differences in the taxonomies that make them non identical even if the taxa is identical
genus_wide <- genus %>% select(-taxa_path) %>% 
  pivot_wider(
    names_from = comparison,
    values_from = -c(names(meta), c(comparison,library_id, taxid, name, rank, lca, PlantAnimal))
  )
```

## Number of reads - genus eukaryotes

### Scatter - total per kingdom 

Let's first plot it like the number of reads at root above
```{r, warning=FALSE}
p <- ggplot(genus_wide, aes(x = nreads_test, y = nreads_stable, color=label)) +
  geom_abline(color = "blue", linetype = "dashed") +
  geom_point(alpha=0.7)+ 
  scale_y_continuous(trans="log10") + 
  scale_x_continuous(trans="log10") + 
  facet_wrap(~PlantAnimal, scales = "free") + 
  labs(
    x = "number of reads (log) - test",
    y = "number of reads (log) - stable",
    title = "Number of reads in eukaryote genera",
    shape = "library type",
    color = "dataset"
  ) + 
  theme_test(base_size = 15) 

ggsave("plots/reads_euk.pdf", plot=p)
print(p)
```

If points fall below the diagonal line, the taxa gets more reads assigned in the test pipeline. If points fall above, the taxa loses reads in the test pipeline. At high values of number of reads (i.e. high values of x and or y), we would want the correlation to be very close to the identity line. A wide spread at low number of reads (i.e. lower left corner), is fine.  

### Scatter - faceted per dataset

```{r, warning=FALSE, message=FALSE}
p <- ggplot(genus_wide, aes(x = nreads_test, y = nreads_stable, color=label)) +
  geom_abline(color = "blue", linetype = "dashed") +
  geom_point(alpha=0.7)+ 
  scale_y_continuous(trans="log10") + 
  scale_x_continuous(trans="log10") + 
  facet_grid2(vars(PlantAnimal),vars(label),scales="free") +
  labs(
    x = "number of reads (log) - test",
    y = "number of reads (log) - stable",
    title = "Number of reads in eukaryote genera - faceted",
    shape = "library type",
    color = "dataset"
  ) + 
  theme_test(base_size = 15) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
ggsave("plots/reads_euk_facet.pdf", plot=p)
print(p)
```

Now the pattern is clearer per dataset. Again, we don't mind that the bottom of the plots is a wide spread, but we need a tight correlation at the top.


### Volcano plot 

Do volcano plot to show if taxa are differently enriched in test/stable. This is a lot of code, explanation at the bottom

```{r, warning=FALSE, message=FALSE}

summary_df <- genus %>%
  group_by(label, PlantAnimal, taxid, name, comparison) %>%
  summarise(total_nreads = sum(nreads), .groups = "drop")

# we filter for taxa with at least 10 reads, otherwise we are just measuring a mess 
wide_df <- summary_df %>%
  filter(total_nreads >= 10) %>% 
  pivot_wider(names_from = comparison, values_from = total_nreads) 

# get the fold change (log2FC)
wide_df <- wide_df %>%
  mutate(log2FC = log2((test + 1) / (stable + 1)))

# p-val per taxa per label
pvals <- genus %>%
  filter(comparison %in% c("test", "stable")) %>%
  group_by(label, PlantAnimal, taxid, name) %>%
  summarise(pval = tryCatch(
    wilcox.test(nreads ~ comparison)$p.value,
    error = function(e) NA_real_
  ), .groups = "drop")

# merge pvalue df in 
volcano_df <- wide_df %>%
  left_join(pvals, by = c("label", "PlantAnimal", "taxid", "name")) %>%
  mutate(log10p = -log10(pval))

# get the average reads per row, just for fancy colours in plotting
volcano_df$avg_nreads <- rowMeans(
  cbind(ifelse(is.na(volcano_df$test), 0, volcano_df$test),
        ifelse(is.na(volcano_df$stable), 0, volcano_df$stable))
)

p <- ggplot(volcano_df, aes(x = log2FC, y = log10p, color = avg_nreads)) +
  geom_point(alpha = 0.7) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed") +
  geom_hline(yintercept = -log10(0.05), linetype = "dotted") +
  labs(
    x = "log 2 fold change (test vs stable)",
    y = "-log 10 (p-value)",
    title = "Volcano plot of taxa",
    subtitle = "only taxa with >= 10 reads",
    color = "mean reads"
  ) +
  scale_colour_viridis_c(trans="log10") +
  facet_grid2(vars(label),vars(PlantAnimal),scales="free") +
  theme_test(base_size = 15) 

ggsave("plots/reads_euk_volcano.pdf", plot=p)
print(p)
```

How to interpret:

  * x-axis goes from (higher reads in stable - equal at 0 - higher reads in test)
  * y axis goes from (not significant - (0.05 threshold dashed line) - significant)
  * the dashed vertical lines show a 2x fold change in either direction

#### Volcano plot - outliers

We can use this data on fold change and significance to find the taxa with the highest fold change between the two runs. 

```{r, warning=FALSE, message=FALSE}
top_hits <- volcano_df %>%
  filter(!is.na(log2FC), !is.na(pval)) %>%
  filter(pval < 0.05) %>% # get significiant
  group_by(label, PlantAnimal) %>%
  slice_max(order_by = abs(log2FC), n = 30, with_ties = FALSE) %>% # keep top 10 taxa per group 
  ungroup() %>%
  mutate(
    name = fct_reorder(name, log2FC)  
  )

for (lab in unique(top_hits$label)) {
  p <- ggplot(top_hits %>% filter(label == lab), aes(x = name, y = log2FC, fill = avg_nreads)) +
  geom_col() +
  facet_wrap(~PlantAnimal, scales = "free") +
  coord_flip() +
  labs(
    title = paste0("top differentially represented taxa in ", lab),
    x = "Taxon",
    y = "log 2 fold change (test vs stable)",
    fill = "mean reads"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text.y = element_text(angle = 0)
  ) + 
  scale_fill_viridis_c(trans="log10") +
  theme_test(base_size = 15) 

  ggsave(paste0("plots/volcano_diff/reads_euk_volcano_diff_", lab, ".pdf"), plot=p) +
  print(p)
}

```


In these, negative values mean we had higher reads in stable, and positive is higher reads in test. The colour is the average number of reads (between both test and stable).

### Taxa gain/loss

That was a fancy way of doing it. But let's just look at how the number of reads changes between the two runs instead
```{r, warning=FALSE, message=FALSE}
# get the difference in nreads
read_diff <- genus %>%
  select(name, taxid, comparison, nreads, PlantAnimal, label) %>%
  group_by(name, taxid, comparison, PlantAnimal, label) %>% 
  summarise(total_nreads = sum(nreads), .groups = "drop") %>%
  pivot_wider(
    names_from = comparison,
    values_from = total_nreads,
    values_fill = 0  # <-- fills NAs directly as 0
  ) %>%
  mutate(read_change = test - stable)

for (l in unique(df$label)) {
  
  d <- read_diff %>%
    filter(label == l) %>%
    group_by(PlantAnimal) %>%
    arrange(desc(abs(read_change))) %>%
    slice_head(n = 30)  # top 30 changes

 p <- ggplot(d, aes(x = reorder(name, read_change), y = read_change, fill = PlantAnimal)) +
  geom_col() +
  coord_flip() +
  labs(
    x = "taxa",
    y = "read count difference (test - stable)",
    title = paste0("Gain/loss of reads per taxa in ", l),
    fill = "kingdom"
  ) +
  theme_test(base_size = 15) 

 t = paste0("plots/nreads_diff/taxa_gainloss_",l,".pdf")
 ggsave(t, plot=p)
 print(p)
}
```
These might be easier to see as pdfs. In these, negative values mean we had higher reads in stable, and positive is higher reads in test. They show the top 30 taxa in the swing in read number. 


## Number of reads total per kingdom

Let's just take each library and sum the genus level reads per kingdom

```{r, warning=FALSE, message=FALSE}

id_df <- genus %>% group_by(library_id, label, PlantAnimal,comparison) %>%
  summarise(total_nreads = sum(nreads)) %>% 
  pivot_wider(names_from = comparison, values_from = total_nreads)

p <- ggplot(id_df, aes(x=test, y =stable, color=label)) +
  geom_abline(color = "blue", linetype = "dashed") +
  geom_point(size = 5) + 
  facet_wrap(~PlantAnimal, scales="free")+ 
  labs(
    x = "total reads - test",
    y = "total reads - stable",
    title = "Total number of reads per kingdom",
    color = "dataset"
  ) +
  theme_test(base_size = 15) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 

ggsave("plots/reads_total_euk.pdf", plot=p)
print(p)  
```

If points fall below the diagonal line, the library gets more reads assigned in the test pipeline. If points fall above, the library loses reads in the test pipeline.


## Number of taxa identified

We can look at the number of taxa identified with at least 1 read
```{r, warning=FALSE, message=FALSE}

id_df <- genus %>% group_by(library_id, label, PlantAnimal,comparison) %>%
  summarise(num = n()) %>% 
  pivot_wider(names_from = comparison, values_from = num)

p <- ggplot(id_df, aes(x=test, y =stable, color=label)) +
  geom_abline(color = "blue", linetype = "dashed") +
  geom_point(size = 5) + 
  facet_wrap(~PlantAnimal, scales="free") +
  labs(
    x = "total number of taxa- test)",
    y = "total number of taxa - stable",
    title = "Total number of taxa per kingdom",
    color = "dataset"
  ) +
  theme_test(base_size = 15) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
ggsave("plots/taxa_euk.pdf", plot=p)
print(p)  
```

If points fall below the diagonal line, the library gets more taxa identified in the test pipeline. If points fall above, the library loses taxa in the test pipeline. Ideally we only see a high divergence here if we are updating the reference catalog. 


## Venn diagrams 

How many taxa are unique to either run? We can do venn diagrams. Here we only include taxa with at least 10 reads per library. 
```{r, warning=FALSE, message=FALSE}

library_ids <- unique(genus$library_id)
plots <- list()

for (lib in library_ids) {
  df_lib <- genus %>%
    filter(library_id == lib & nreads >= 10)

  label <- unique(df_lib$label)

  venn_plots <- list()
  for (group in unique(df_lib$PlantAnimal)) {
    df_group <- df_lib %>%
      filter(PlantAnimal == group)

    taxa_old <- df_group %>%
      filter(comparison == "stable") %>%
      pull(name) %>%
      unique()

    taxa_new <- df_group %>%
      filter(comparison == "test") %>%
      pull(name) %>%
      unique()

    venn_counts <- list(
      old = taxa_old,
      new = taxa_new
    )

    fit <- tryCatch(euler(venn_counts), error = function(e) NULL)

    if (!is.null(fit)) {
      p <- plot(
        fit,
        fills = list(fill = c("royalblue4", "tomato4"), alpha = 0.5),
        quantities = TRUE,
        main = paste("Library:", lib, "|", group)
      )
      venn_plots <- append(venn_plots, list(p))
    }
  }

  if (length(venn_plots) > 0) {
    combined <- do.call(gridExtra::arrangeGrob, c(venn_plots, ncol = 2,
      top = paste("Taxa (>=10 reads) overlap for library:", lib, "-", label)))
    plots <- append(plots, list(combined))
  }
}

pdf("plots/venn_taxa_by_library.pdf", width = 10, height = 8)
for (p in plots) {
  grid::grid.draw(p)
  grid::grid.newpage()
}
dev.off()

```

The venn diagrams are output to a pdf because they are big. Look there. Unless we changed the reference catalog, we hope for a big overlap in the sets. 


## Top unique taxa 

Let's look at what taxa were unique. 

```{r, warning=FALSE, message=FALSE}
top_n <- 20 # take 20 top
plots <- list()

for (lab in unique(genus$label)) {
  
  df_lab <- genus %>%
    filter(label == lab, nreads >= 10) # at least 10 reads
  
  # get ones that only appear in test or stable
  unique_taxa <- df_lab %>%
    group_by(name, taxid, PlantAnimal) %>%
    filter(n_distinct(comparison) == 1) %>%
    ungroup()
  
  # summarise to get total reads, and find the top n
  top_taxa <- unique_taxa %>%
    group_by(name, comparison, PlantAnimal) %>%
    summarise(total_reads = sum(nreads), .groups = "drop") %>%
    group_by(comparison, PlantAnimal) %>%
    slice_max(total_reads, n = top_n, with_ties = FALSE) %>%
    ungroup() %>%
    mutate(name = fct_reorder(name, total_reads))
  
  # skip if there is nothing 
  if (nrow(top_taxa) == 0) next
  
  p <- ggplot(top_taxa, aes(x = name, y = total_reads, fill = comparison)) +
    geom_col(position = "dodge") +
    coord_flip() +
    facet_wrap(~PlantAnimal, scales = "free") +
    labs(
      title = paste0("Top ", top_n, " Unique Taxa in ", lab),
      x = "taxon",
      y = "total reads",
      fill = "comparison"
    ) + theme_test(base_size = 15) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 

  
  fname <- paste0("plots/unique_taxa/top_unique_taxa_", lab, ".pdf")
  ggsave(fname, plot = p, width = 10, height = 7)
  print(p)
}

```

Here taxa unique to stable and test (the colours) are shown per dataset with their total amount of reads. We should look to see if any of these unique taxa are specifically informative or otherwise worrying. 


## Top taxa in new version (test)

Now let's just look at what taxa (unique or not) dominate the test run.

```{r, warning=FALSE, message=FALSE}
top_n <- 20 # show top 20 taxa
for (lab in unique(genus$label)) {
  
  df_lab <- genus %>%
    filter(label == lab, comparison == "test", nreads >= 10) # at least 10 reads
  
  top_taxa <- df_lab %>%
    group_by(name, PlantAnimal) %>%
    summarise(total_reads = sum(nreads), .groups = "drop") %>%
    group_by(PlantAnimal) %>%
    slice_max(total_reads, n = top_n, with_ties = FALSE) %>%
    ungroup() %>%
    mutate(name = fct_reorder(name, total_reads))
  
  if (nrow(top_taxa) == 0) next
  
  p <- ggplot(top_taxa, aes(x = name, y = total_reads, fill = PlantAnimal)) +
    geom_col() +
    coord_flip() +
    facet_wrap(~PlantAnimal, scales = "free") +
    labs(
      title = paste0("Top ", top_n, " Unique Taxa in ", lab),
      x = "taxon",
      y = "total reads",
      fill = "comparison"
    ) + theme_test(base_size = 15) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
  
  fname <- paste0("plots/top_taxa/top_taxa_test_", lab, ".pdf")
  ggsave(fname, plot = p, width = 10, height = 7)
  print(p)
}

```

This now just shows the most abundant taxa in each dataset if we use the pipeline being tested. 

## Heatmaps 

### Heatmap of test pipeline 

Let's just do a heatmap showing the most abudant taxa per dataset, now we can see per library too 

```{r}
top_n <- 100  # top n

df_genus <- genus %>%
  filter(comparison == "test" & nreads >= 10, !is.na(name)) %>%
  mutate(taxon = paste0(name, "-", taxid))

for (lab in unique(df_genus$label)) {
  
  df_lab <- df_genus %>%
    filter(label == lab)
  
  for (group in c("plant", "animal")) {
    
    df_group <- df_lab %>%
      filter(PlantAnimal == group)
    
    if (nrow(df_group) == 0) next
    
    # top n taxa by abundance
    top_taxa <- df_group %>%
      group_by(taxon) %>%
      summarise(total = sum(nreads), .groups = "drop") %>%
      slice_max(total, n = top_n, with_ties = FALSE) %>%
      pull(taxon)
    
    df_top <- df_group %>%
      filter(taxon %in% top_taxa)
    
    # aggregate read counts per taxon per library
    df_plot <- df_top %>%
      group_by(library_id, taxon) %>%
      summarise(total_reads = sum(nreads), .groups = "drop") %>%
      mutate(log_reads = log10(total_reads + 1))
    
    # order taxa and libraries for readability
    taxon_order <- df_plot %>%
      group_by(taxon) %>%
      summarise(total = sum(total_reads)) %>%
      arrange(desc(total)) %>%
      pull(taxon)
    
    lib_order <- df_lab %>%
      distinct(library_id) %>%
      pull(library_id)
    
    df_plot$taxon <- factor(df_plot$taxon, levels = taxon_order)
    df_plot$library_id <- factor(df_plot$library_id, levels = lib_order)
    
    # plot
    p <- ggplot(df_plot, aes(x = library_id, y = taxon, fill = log_reads)) +
      geom_tile() +
      scale_fill_viridis_c(name = "log10(reads + 1)") +
      labs(
        title = paste("Top", top_n, "Genera in", lab, "(", group, ")"),
        x = "library ID",
        y = "taxon"
      ) +
      theme_test(base_size = 15) +
      theme(
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        panel.grid = element_blank()
      )
    
    # Save
    fname <- paste0("plots/heatmaps/test_heatmap_top100_", lab, "_", group, ".pdf")
    ggsave(fname, plot = p, width = 10, height = 14)
    print(p)
  }
}


```

These are big heatmaps, more easily viewed in the pdf version. We can see if all libraries of a dataset share similar taxa. 

### Ratio heatmap 

Let's do a heatmap where we can look at the difference between test and stable
```{r}
top_n <- 10  # top taxa to include per group

df_genus <- genus %>%
  filter(nreads >= 100, !is.na(name)) %>%
  mutate(taxon = paste0(name, "-", taxid))

# get top taxa per group (across test+stable)
top_taxa <- df_genus %>%
  group_by(taxon, PlantAnimal) %>%
  summarise(total = sum(nreads), .groups = "drop") %>%
  group_by(PlantAnimal) %>%
  slice_max(total, n = top_n, with_ties = FALSE) %>%
  pull(taxon)

df_top <- df_genus %>%
  filter(taxon %in% top_taxa)

# sum reads per taxon/library/comparison
df_sum <- df_top %>%
  group_by(label, PlantAnimal, library_id, taxon, comparison) %>%
  summarise(total_reads = sum(nreads), .groups = "drop")

# to wide
df_wide <- df_sum %>%
  pivot_wider(
    names_from = comparison,
    values_from = total_reads,
    values_fill = 0  # replace missing values with 0
  ) %>%
  mutate(
    log2_ratio = log2((test + 1) / (stable + 1))
  )

# order
df_wide <- df_wide %>%
  group_by(taxon) %>%
  mutate(taxon_total = sum(test + stable)) %>%
  ungroup()

taxon_order <- df_wide %>%
  group_by(taxon) %>%
  summarise(order_val = sum(abs(log2_ratio), na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(order_val)) %>%
  pull(taxon)

for (lab in unique(df_wide$label)) {
  for (group in c("plant", "animal")) {
    
    df_plot <- df_wide %>%
      filter(label == lab, PlantAnimal == group)
    
    if (nrow(df_plot) == 0) next
    
    df_plot$taxon <- factor(df_plot$taxon, levels = taxon_order)
    df_plot$library_id <- factor(df_plot$library_id)

    p <- ggplot(df_plot, aes(x = library_id, y = taxon, fill = log2_ratio)) +
      geom_tile() +
      scale_fill_gradient2(
        name = "log2(test / stable)",
        low = "blue", high = "red", mid = "white",
        midpoint = 0, limits = c(-4, 4), oob = scales::squish
      ) +
      labs(
        title = paste("Log2 ratio heatmap (test vs stable)", "\ndataset:", lab, "| Group:", group),
        x = "library ID",
        y = "taxon"
      ) +
      theme_test(base_size = 15) +
      theme(
        axis.text.x = element_text(angle = 90, hjust = 1),
        panel.grid = element_blank()
      )
    
    ggsave(paste0("plots/heatmaps/heatmap_log2ratio_", lab, "_", group, ".pdf"), plot = p, width = 10, height = 14)
    print(p)
  }
}
```

These new heatmaps now show the difference between test and stable. Red taxa have higher reads in test, and blue had higher reads in stable.


## PCoA

Let's do PCoA. It's like PCA but with a little o. 

### Regular 

Do PCoA, three plots per label (plants, animals, fungi), each dot is one library, lines go between the dots, length of line is how dissimilar they are

```{r, message=FALSE, eval = FALSE}

plots <- list()

for (lab in unique(genus$label)) {

  for (group in c("plant", "animal", "fungi")) {

    genus_sub <- genus %>%
      filter(label == lab, PlantAnimal == group, !is.na(name), !is.na(nreads)) %>%
      mutate(
        taxa = paste0(name, "-", taxid),
        sample = paste(library_id, comparison, sep = "_")
      )


    # wide
    tmp_dt <- as.data.table(genus_sub)
    wide_counts <- dcast(tmp_dt, taxa ~ sample, value.var = "nreads", fill = 0)

    if (nrow(wide_counts) < 2) next

    rownames(wide_counts) <- wide_counts$taxa
    wide_data <- as.data.frame(wide_counts[, -1])

    # compute distances
    dist_matrix <- vegdist(t(wide_data), method = "bray")
    pcoa_result <- cmdscale(dist_matrix, k = 2, eig = TRUE)

    pcoa_df <- data.frame(pcoa_result$points)
    colnames(pcoa_df) <- c("PCoA1", "PCoA2")
    pcoa_df$sample <- rownames(pcoa_result$points)

    # join metadata
    pcoa_df <- pcoa_df %>%
      separate(sample, into = c("library_id", "comparison"), sep = "_", remove = FALSE) %>%
      left_join(
        genus_sub %>% 
          select(library_id, comparison) %>%
          distinct(),
        by = c("library_id", "comparison")
      )

    # plot
    p <- ggplot(pcoa_df, aes(x = PCoA1, y = PCoA2, color = comparison)) +
      geom_path(aes(group = library_id), color = "grey50", linewidth = 0.5, alpha = 0.7) +
      geom_point(size = 5) +
      labs(
        title = paste("PCoA | Label:", lab, "| Group:", group),
        x = "PCoA1",
        y = "PCoA2",
        color = "Pipeline"
      ) +
      theme_test(base_size = 15)

    plots <- append(plots, list(p))
  }
}

pdf("plots/pcoa.pdf", width = 10, height = 8)
for (p in plots) print(p)
dev.off()


```

Look at the pdf to see this PCoAs. Each library becomes a point. The distance between libraries in test and stable is joined by a line. The bigger the line, the greater the difference. 

### Exclusion

Can we do a thing where we remove taxa one by one, and measure if the points become nearer? That would find the taxa that drive the difference between runs. Turns out yes, we can 

```{r, eval = FALSE}

plots <- list()

for (lab in unique(genus$label)) {

  for (group in c("plant", "animal", "fungi")) {
    
    df2 <- genus %>%
      filter(label == lab, PlantAnimal == group, !is.na(name), !is.na(nreads)) %>%
      mutate(
        taxa = paste0(name, "-", taxid),
        sample = paste(library_id, comparison, sep = "_")
      )
    
    tmp_dt <- as.data.table(df2)
    wide_counts <- dcast(tmp_dt, taxa ~ sample, value.var = "nreads", fill = 0)
    rownames(wide_counts) <- wide_counts$taxa
    wide_data <- as.data.frame(wide_counts[, -1])
    rownames(wide_data) <- wide_counts$taxa

    # get the full PCoA distance between old/new for each library
    dist_matrix <- vegdist(t(wide_data), method = "bray")
    pcoa <- cmdscale(dist_matrix, k = 2)
    coord_full <- data.frame(pcoa)
    coord_full$sample <- rownames(coord_full)

    # extract library_id + comparison
    coord_full <- coord_full %>%
      separate(sample, into = c("library_id", "comparison"), sep = "_", remove = FALSE)
    
    # get distance between old and new per library
    baseline_diffs <- coord_full %>%
      pivot_wider(id_cols = library_id, names_from = comparison, values_from = c(X1, X2)) %>%
      mutate(
        euclidean = sqrt((X1_stable - X1_test)^2 + (X2_stable - X2_test)^2)
      ) %>%
      select(library_id, baseline = euclidean)

    taxa_list <- rownames(wide_data)
    results <- list()
    
    for (tax in taxa_list) {
      # remove taxon row
      reduced_data <- wide_data[setdiff(rownames(wide_data), tax), ]
    
      dist_reduced <- vegdist(t(reduced_data), method = "bray")
      pcoa_reduced <- cmdscale(dist_reduced, k = 2)
    
      coords <- data.frame(pcoa_reduced)
      coords$sample <- rownames(coords)
      coords <- coords %>%
        separate(sample, into = c("library_id", "comparison"), sep = "_", remove = FALSE)
    
      diffs <- coords %>%
        pivot_wider(id_cols = library_id, names_from = comparison, values_from = c(X1, X2)) %>%
        mutate(
          euclidean = sqrt((X1_stable - X1_test)^2 + (X2_stable - X2_test)^2)
        ) %>%
        select(library_id, euclidean) %>%
        left_join(baseline_diffs, by = "library_id") %>%
        mutate(taxa_removed = tax, delta = baseline - euclidean)
    
      results[[tax]] <- diffs
    }
    
    # get average nreads
    abundance_df <- df2 %>%
      group_by(taxa) %>%
      summarise(avg_reads = mean(nreads, na.rm = TRUE), .groups = "drop")

  
    influence_df <- bind_rows(results)
    
    # summarise: total distance reduction across libraries
    summary_df <- influence_df %>%
      group_by(taxa_removed) %>%
      summarise(mean_delta = mean(delta, na.rm = TRUE), .groups = "drop") %>%
      left_join(abundance_df, by = c("taxa_removed" = "taxa")) %>%
      arrange(desc(mean_delta))

    filtered_df <- summary_df %>%
      arrange(mean_delta) %>%
      slice(c(1:10, (n() - 9):n()))
    
     p <- ggplot(filtered_df, aes(x = reorder(taxa_removed, mean_delta), y = mean_delta, fill = avg_reads)) +
      geom_col() +
      coord_flip() +
      scale_fill_viridis_c(name = "Mean Reads") +
      labs(
        title = paste("PCoA influencers | Label:", lab, "| Group:", group),
        x = "taxa (genus-taxid)",
        y = "mean difference in PCoA",
        fill = "mean reads"
      ) +
      theme_test(base_size = 15)
    plots <- append(plots, list(p))
  }
}


pdf("plots/pcoa_influence.pdf", width = 10, height = 8)
for (p in plots) print(p)
dev.off()

```

Look in the pdf for these plots. Positive values mean that this taxa is *driving* the separation of the test and stable run in PCoA space. Negative values mean the taxa is bringing the points closer together. 


## Damage plots 

We are obviously interested in the damage per taxa

### Big pdf 

This is a function to plot damage patterns. It is complicated to do apparently. 
```{r}

plot_damage <- function(df2, only5 = FALSE) {

  # Get info
  nreads_old <- df2 %>% filter(comparison == "stable") %>% pull(nreads)
  nreads_new <- df2 %>% filter(comparison == "test") %>% pull(nreads)
  A_b_old <- df2 %>% filter(comparison == "stable") %>% pull(A)
  A_b_new <- df2 %>% filter(comparison == "test") %>% pull(A)
  lib_id <- df2$library_id[1]
  label <- df2$label[1]
  taxon <- df2$name[1]
  taxid <- df2$taxid[1]

  x_cols <- grep("^fwdx\\d+$", names(df2), value = TRUE)
  f_cols <- grep("^fwf\\d+$", names(df2), value = TRUE)

  long_df <- bind_rows(
    df2 %>%
      filter(comparison == "stable") %>%
      select(all_of(x_cols), all_of(f_cols)) %>%
      pivot_longer(everything(), names_to = "metric", values_to = "value") %>%
      mutate(position = as.integer(str_extract(metric, "\\d+")),
             type = ifelse(str_starts(metric, "fwdx"), "x", "f"),
             comparison = "stable"),
    df2 %>%
      filter(comparison == "test") %>%
      select(all_of(x_cols), all_of(f_cols)) %>%
      pivot_longer(everything(), names_to = "metric", values_to = "value") %>%
      mutate(position = as.integer(str_extract(metric, "\\d+")),
             type = ifelse(str_starts(metric, "fwdx"), "x", "f"),
             comparison = "test")
  ) %>%
    filter(!is.na(value)) %>%
    mutate(type = factor(type, levels = c("f", "x")))

  if (only5) {
    long_df <- long_df %>% filter(position < 7)
  }

  A_points <- tibble(
    position = 0,
    value = c(A_b_old, A_b_new),
    type = "A",
    comparison = c("stable", "test")
  )

  title <- paste0(
    taxon, " (taxid: ", taxid, ")\n",
    "Library: ", lib_id, ", Label: ", label, "\n",
    "nreads (stable/test): ", nreads_old, " / ", nreads_new,
    ", A: ", round(A_b_old, 3), " / ", round(A_b_new, 3)
  )
  ggplot(long_df, aes(x = position, y = value, color = comparison, linetype = type)) +
    geom_line() +
    geom_point(data = filter(long_df, type == "f")) +
    geom_point(data = A_points, aes(shape = type), size = 3, stroke = 1.2, inherit.aes = TRUE) +
    scale_color_manual(values = c("stable" = "royalblue4", "test" = "tomato4")) +
    scale_linetype_manual(values = c("x" = "solid", "f" = "dashed")) +
    scale_shape_manual(values = c(A = 8)) +
    labs(
      title = title,
      x = "position from 5prime end",
      y = "C-T",
      color = "comparison",
      linetype = "metric",
      shape = NULL
    ) +
     theme_test(base_size = 15)
}

```

We can use the function we just made to make damage plots for all taxa with at least 100 reads in test and stable
```{r, warning=FALSE, message=FALSE, eval = FALSE}
# get taxa with >= 100 reads
valid_taxa <- genus %>%
  group_by(name, taxid, library_id) %>%
  summarise(
    n_old = nreads[comparison == "stable"],
    n_new = nreads[comparison == "test"],
    .groups = "drop"
  ) %>%
  filter(n_old >= 100, n_new >= 100)

genus_filtered <- semi_join(genus, valid_taxa, by = c("name", "taxid", "library_id"))

# split into groups 
groups <- genus_filtered %>%
  group_by(name, taxid, library_id) %>%
  group_split()

# save plots 
plots <- list()
for (df2 in groups) {
  p <- plot_damage(df2)
    plots <- append(plots, list(p))
}

pdf("plots/damage_patterns.pdf", width = 8, height = 6)
for (p in plots) {
  print(p)
}
dev.off()


```

Look at the pdf that is full of these. The blue lines show stable, and red show test. The dotted lines show the exact measured C-T, and the smoothed line is the estimate. The dfit (A, position 1 C-T corrected for background subsitution rate) is shown by a star. We want the damage patterns to look similar to each other. 

### Just damage estimates (A)

```{r}
p <- genus_wide %>% filter(nreads_test >= 100 & nreads_stable >= 100) %>%
ggplot(aes(x=A_test, y = A_stable, color = label, size = nreads_stable)) +
  geom_abline(color = "blue", linetype = "dashed") +
  geom_point(alpha = 0.7) +
  facet_wrap(~PlantAnimal, scales = "free") + 
  labs(
    x = "A - test",
    y = "A - stable",
    title = "DNA damage (A) in eukaryotes",
    color = "dataset", 
    size = "reads in stable"
  ) + 
  theme_test(base_size = 15)
ggsave("plots/dmg_euk.pdf", plot=p)
print(p)  
```


This shows if any taxa have a big difference in damage profiles. This is just taxa with at least 100 reads in both runs so should be stable. Sometimes it is just metaDMG that drives differences here. 

# Taxa shared by datasets

## All datasets 

Taxa shared by many datasets could indicate a huge contamination or just a bad reference genome. Which taxa are shared by all datasets?

```{r, warning=FALSE, message=FALSE}
all_labels <- unique(genus$label)
n_total_labels <- length(all_labels)
abundant_taxa <- genus %>%
  filter(nreads >= 50)

ubiquitous_taxa <- abundant_taxa %>%
  group_by(name, PlantAnimal, comparison) %>%
  summarise(
    n_labels = n_distinct(label),
    total_reads = sum(nreads),
    .groups = "drop"
  ) %>%
  filter(n_labels == n_total_labels)

p <- ggplot(ubiquitous_taxa, aes(x = reorder(name, total_reads), y = total_reads, fill = PlantAnimal)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Taxa found in all datasets (≥50 reads)",
    x = "taxon",
    y = "total reads"
  ) + facet_wrap(~comparison) + scale_y_continuous(trans="log10") + 
    theme_test(base_size = 15)
ggsave("plots/ubiquitous_taxa.pdf" , plot = p, width = 10, height = 7)
print(p)

```

Look at these taxa. Are any super abundant? Are any super important for conclusions?

### All datasets - damage

Find damage of these bad taxa?

```{r, warning=FALSE, message=FALSE}
bad <- genus %>% filter(name %in% ubiquitous_taxa$name)

p <- ggplot(bad, aes(x=nreads,y=A,shape=comparison,color=name)) + 
  geom_point() + scale_x_continuous(trans="log10") + 
   labs(
    title = "Damage in taxa found in all datasets (>=50 reads)",
    x = "number of reads",
    y = "DNA damage (A)"
  ) + 
   theme_test(base_size = 15)
ggsave("plots/ubiquitous_taxa_dmg.pdf" , plot = p, width = 10, height = 7)
print(p)
```

Are there any very damaged taxa? 


## At least 4 datasets

Which are shared by at least 4?

```{r}
abundant_taxa <- genus %>%
  filter(nreads >= 50)

# count how many groups each taxa is in 
taxa_presence <- abundant_taxa %>%
  group_by(name, PlantAnimal, comparison) %>%
  summarise(
    n_labels = n_distinct(label),
    total_reads = sum(nreads),
    .groups = "drop"
  )

# filter for taxa in at least 4
taxa_filtered <- taxa_presence %>%
  filter(n_labels >= 4)

for (pl in unique(taxa_filtered$PlantAnimal)) {
  for (s in c("stable","test")) {
    

  df_sub <- taxa_filtered %>% filter(PlantAnimal == pl & comparison == s)

  p <- ggplot(df_sub, aes(x = reorder(name, total_reads), y = n_labels, fill = total_reads)) +
    geom_col(position = position_dodge()) +
    coord_flip() +
    labs(
      title = paste0(pl, " taxa appearing across multiple datasets in ", s),
      x = "taxa",
      y = "# datasets",
      fill = "number of reads"
    ) + facet_wrap(~comparison)

  fname <- paste0("plots/common_taxa/common_taxa_", pl,"_",s, ".pdf")
  ggsave(fname, plot = p, width = 10, height = 7)
  print(p)
  }
}


```


These plots show the taxa that appear in at least 4 datasets (per kingdom and per comparison)

# LCA file 

We want to see how many reads are classified at each rank. For this we need the lca.gz files. We need to pre-process these on the server end because they are huge

```{bash, eval = FALSE}
# find lca files 
for line in $(find /projects/caeg/data/validation -path "*/results/metadmg/lca/*" -name '*_collapsed.lca.gz' | grep 'v1.0.3')
do   
  status=""
  if echo "$line" | grep -q "eebd93d8414a08f40337174bb00ee3dc"; then # if this is in path, that means it's the new catalog
      status="test"
  fi
  if echo "$line" | grep -q "48ca4558f97def8149639ed5c1b40c37"; then # this is v1
      status="stable"
  fi
  if [ ! -z $status ]; then 
    echo ln -s "$line" "validation_links/${status}_$(basename $line)" 
  fi 
done 

# make a tsv files with nreads per rank 
for f in $(ls validation_links/*lca.gz)
do
  for kingdom in Metazoa Viridiplantae Fungi
  do 
    echo "zcat $f|grep $kingdom|cut -f6|rev|cut -f1 -d":"|rev|sort|uniq -c > tmp/$(basename $f).$kingdom"
  done 
done 

```

Now we load all those small files into a big df 
```{R, eval = FALSE}
library(tidyverse)
library(data.table)

file_list <- list.files(
  path = "tmp/",
  recursive = FALSE,
  full.names = TRUE
)

read_file <- function(f) {
  print(f)
  df <- fread(f, header=F, fill=T)
  df$sample_id <- f
  return(df)
}

# read the files into a big df 
df <- do.call(rbind, lapply(file_list, read_file))

write.csv(df,"lca.csv",row.names=F)
```

And download it 
```{bash, eval = FALSE}
rsync -av dlm551@dandycomp01fl:/maps/projects/caeg/people/dlm551/ValidationHelpers/lca.csv .
```

Now we can load it in and plot it
```{r}
lca <- fread("lca.csv") 
names(lca) <- c("nreads","rank","filename")
lca <- lca %>%
  mutate(
    base = basename(filename),
    comparison = sapply(strsplit(base, "_"), `[`, 1),
    library_id  = sapply(strsplit(base, "_"), `[`, 3),
    kingdom  = sapply(strsplit(base, "\\."), `[`, 4),
  ) %>%
  select(-base)


lca_wide <- lca %>%
  pivot_wider( 
    id_cols = c(library_id, rank, kingdom),
    names_from = comparison,
    values_from = nreads,
    values_fill = list(nreads = 0)
  )

lca_wide <- lca_wide %>% left_join(meta, by = "library_id")

```

Plot just the major ranks 
```{r, warning=FALSE, message=FALSE}

major_ranks <- c("species", "genus", "family", "order", "class", "phylum", "kingdom", "superkingdom")

for (l in unique(lca_wide$label)) {
  
   this_lca <- lca_wide %>%
    filter(rank %in% major_ranks & label == l) %>%
    mutate(rank = factor(rank, levels = major_ranks))

  p <- ggplot(this_lca, aes(x=test, y = stable, color = rank)) +
    geom_abline(color = "blue", linetype = "dashed") +
    geom_point(size = 5) + 
    facet_wrap(~kingdom,scales="free") +
    labs(
    x = "number of reads - test",
    y = "number of reads- stable",
    title = paste0("Number of reads per rank in ", l),
    fill = "rank"
    )
    theme_test(base_size = 15) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
  
   t = paste0("plots/lca/lca_",l,".pdf")
   ggsave(t, plot=p)
   print(p)
}

```

In these plots, in points are below the line it means that the rank is more common in test than in stable. 



```{r}
```

```{r}
```

```{r}
```

```{r}
```
