---
title: "Validation"
output: html_document
date: "2025-07-08"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```




This relies on some bash magic that we used server side to rename the files before download, so that the name is e.g. OLD_Lib_LV123_collapsed.stat.gz, and symlinks the results to a folder which can then be downloaded using rsync. 

```{bash eval=FALSE}
#ssh dandycomp01fl
cd /projects/caeg/data/validation
for line in $(find Kap LV? -name '*_collapsed.stat.gz' | grep 'v1.0.3' | grep 'results/metadmg/aggregate')
    if echo "$line" | grep -q "eebd93d8414a08f40337174bb00ee3dc"; then # if this is in path, that means it's the new catalog
        status="new"
    else
        status="old"
    fi
    basename=$(basename "$line")
   ln -s "/projects/caeg/data/validation/$line" "/home/dlm551/valfol/${status}_$basename" 
done 
```

Start with libraries
```{r}
library(tidyverse)
library(data.table)
library(dtplyr)
library(rlang)
library(ggforce)
library(wesanderson)
library(viridis)
library(ggh4x)
library(vegan)
library(ggrepel)
```

We need all the aggregated stat files
```{r}
read_file <- function(f) {
  df <- fread(f, header=T, sep="\t", fill=T, nThread=2)
  df$sample_id <- f
  return(df)
}

file_list <- list.files(
  path = "aggregate",
  pattern = "\\.gz$",  
  recursive = FALSE,
  full.names = TRUE
)

df <- do.call(rbind, lapply(file_list, read_file))
```

Now get library ID and pipeline version from the filename. This is the final df, we can do some funky plots with it
```{r}
df <- df %>%
  mutate(
    base = basename(sample_id),
    pipeline = sapply(strsplit(base, "_"), `[`, 1),
    library  = sapply(strsplit(base, "_"), `[`, 3)
  ) %>%
  select(-base)
```

Get a df with just the root stats and make wide for easy plotting
```{r}
root <- df %>% filter(name == "root")
root_wide <- root %>%
  pivot_wider(
    id_cols = c(library, taxid, name, rank, lca, taxa_path),
    names_from = pipeline,
    values_from = -c(library, pipeline, taxid, name, rank, lca, taxa_path)
  )
```

Number of reads assigned total 
```{r}
ggplot(root_wide, aes(x = nreads_new, y = nreads_old)) +
  # 10% band around y = x
  geom_ribbon(aes(ymin = nreads_new * 0.9, ymax = nreads_new * 1.1), 
              fill = "grey80", alpha = 0.5) +
  # y = x line
  geom_abline(color = "blue", linetype = "dashed") +
  # points
  geom_point() +
  # labels for points outside ±10%
  geom_text_repel(
    data = root_wide %>% filter(abs(nreads_old - nreads_new) > 0.1 * nreads_new),
    aes(label = library), 
    size = 3
  ) +
  labs(
    x = "nreads (new pipeline)",
    y = "nreads (old pipeline)",
    title = "Comparison of nreads between pipelines",
    subtitle = "Shaded band = ±10% difference from identity line"
  )

```


Average read length
```{r}
ggplot(root_wide, aes(x = mean_rlen_new, y = mean_rlen_old)) +
  # 10% band around y = x
  geom_ribbon(aes(ymin = mean_rlen_new * 0.9, ymax = mean_rlen_new * 1.1), 
              fill = "grey80", alpha = 0.5) +
  # y = x line
  geom_abline(color = "blue", linetype = "dashed") +
  # points
  geom_point(aes(color=nreads_new)) +
  scale_color_viridis(trans = "log10", option = "plasma", direction = 1) +
  # labels for points outside ±10%
  geom_text_repel(
    data = root_wide %>% filter(abs(mean_rlen_old - mean_rlen_new) > 0.1 * mean_rlen_new),
    aes(label = library), 
    size = 3
  ) +
  labs(
    x = "rlen (new pipeline)",
    y = "rlen (old pipeline)",
    title = "Comparison of read length between pipelines",
    subtitle = "Shaded band = ±10% difference from identity line"
  )
```


Average damage
```{r}
ggplot(root_wide, aes(x = A_new, y = A_old)) +
  # 10% band around y = x
  geom_ribbon(aes(ymin = A_new * 0.9, ymax = A_new * 1.1), 
              fill = "grey80", alpha = 0.5) +
  # y = x line
  geom_abline(color = "blue", linetype = "dashed") +
  # points
  geom_point(aes(color=nreads_new)) +
  scale_color_viridis(trans = "log10", option = "plasma", direction = 1) +
  # labels for points outside ±10%
  geom_text_repel(
    data = root_wide %>% filter(abs(A_old - A_new) > 0.1 * A_new),
    aes(label = library), 
    size = 3
  ) +
  labs(
    x = "A (new pipeline)",
    y = "A (old pipeline)",
    title = "Comparison of damage (A) between pipelines",
    subtitle = "Shaded band = ±10% difference from identity line"
  )
```


Get just genus level, only eukaryotes and name the kingdoms
```{r}
genus <- df %>% filter(rank == "genus") |> # must be classified at genus level 
    filter(grepl("Eukaryota", taxa_path)) |> # must be a Eukaryote 
    mutate(PlantAnimal = case_when( # define plant/animal based on taxpath 
      grepl("Viridiplantae", taxa_path) ~ "plant",
      grepl("Metazoa", taxa_path) ~ "animal",
      grepl("Fungi", taxa_path) ~ "fungi")) %>% 
  filter(!is.na(PlantAnimal))
      
```



Do long to wide with genus df, here we remove taxa path because it is slightly different between the two taxonomy versions
```{r}
genus_wide <- genus %>% select(-taxa_path) %>% 
  pivot_wider(
    id_cols = c(library, taxid, name, rank, lca,rank,PlantAnimal),
    names_from = pipeline,
    values_from = -c(library, pipeline, taxid, name, rank, lca,rank,PlantAnimal)
  )
```

```{r}
ggplot(genus_wide, aes(x = nreads_new, y = nreads_old, color=PlantAnimal)) +
  geom_abline() + 
  geom_point()+ 
  scale_y_continuous(trans="log10") + 
  scale_x_continuous(trans="log10")
```

```{r}
```

```{r}
```