---
title: "Validation"
output: html_document
date: "2025-07-08"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Setup

This relies on some bash magic that we used server side to rename the files before download, so that the name is e.g. OLD_Lib_LV123_collapsed.stat.gz, and symlinks the results to a folder which can then be downloaded using rsync. 

```{bash eval=FALSE}
#ssh dandycomp01fl
cd /projects/caeg/data/validation
for line in $(find Kap LV? -name '*_collapsed.stat.gz' | grep 'v1.0.3' | grep 'results/metadmg/aggregate')
    if echo "$line" | grep -q "eebd93d8414a08f40337174bb00ee3dc"; then # if this is in path, that means it's the new catalog
        status="new"
    else
        status="old"
    fi
    basename=$(basename "$line")
   ln -s "/projects/caeg/data/validation/$line" "/home/dlm551/valfol/${status}_$basename" 
done 
```

We also need to download the metadata. Download from here https://docs.google.com/spreadsheets/d/16m6KTU2PfVh_40NzRZMKmTlEiArrnWHE4dMbYiGF4n8/edit?gid=0#gid=0 and save as metadata.csv 

# Read in data 

Start with libraries
```{r}
library(tidyverse)
library(data.table)
library(dtplyr)
library(rlang)
library(ggforce)
library(wesanderson)
library(viridis)
library(ggh4x)
library(vegan)
library(ggrepel)
library(purrr)
library(eulerr)
```

Get the metadata 
```{r}
meta <- fread("metadata.csv")
```


We need all the aggregated stat files
```{r}
read_file <- function(f) {
  df <- fread(f, header=T, sep="\t", fill=T, nThread=2)
  df$sample_id <- f
  return(df)
}

file_list <- list.files(
  path = "aggregate",
  pattern = "\\.gz$",  
  recursive = FALSE,
  full.names = TRUE
)

df <- do.call(rbind, lapply(file_list, read_file))
```

Now get library ID and comparison from the filename. This is the final df, we can do some funky plots with it
```{r}
df <- df %>%
  mutate(
    base = basename(sample_id),
    comparison = sapply(strsplit(base, "_"), `[`, 1),
    library_id  = sapply(strsplit(base, "_"), `[`, 3)
  ) %>%
  select(-base)
```

```{r}
df <- df %>% left_join(meta)
```
# Root plots 

Get a df with just the root stats and make wide for easy plotting
```{r}
root <- df %>% filter(name == "root")
root_wide <- root %>%
  pivot_wider(
    id_cols = c(names(meta), c(library_id, taxid, name, rank, lca, taxa_path)),
    names_from = comparison,
    values_from = -c(names(meta), c(comparison,library_id, taxid, name, rank, lca, taxa_path))
  )
```

Number of reads assigned total 
```{r}
p <- ggplot(root_wide, aes(x = nreads_new, y = nreads_old)) +
  # 10% band around y = x
  geom_ribbon(aes(ymin = nreads_new * 0.9, ymax = nreads_new * 1.1), 
              fill = "grey80", alpha = 0.5) +
  # y = x line
  geom_abline(color = "blue", linetype = "dashed") +
  # points
  geom_point(aes(color=label, shape=library_type)) +
  # labels for points outside ±10%
  geom_text_repel(
    data = root_wide %>% filter(abs(nreads_old - nreads_new) > 0.1 * nreads_new),
    aes(label = library_id), 
    size = 3
  ) +
  labs(
    x = "nreads (new comparison)",
    y = "nreads (old comparison)",
    title = "Comparison of nreads between comparisons",
    subtitle = "Shaded band = ±10% difference from identity line"
  )
ggsave("plots/reads_root.pdf", plot=p)
print(p)
```


Average read length
```{r}
p <- ggplot(root_wide, aes(x = mean_rlen_new, y = mean_rlen_old)) +
  # 10% band around y = x
  geom_ribbon(aes(ymin = mean_rlen_new * 0.9, ymax = mean_rlen_new * 1.1), 
              fill = "grey80", alpha = 0.5) +
  # y = x line
  geom_abline(color = "blue", linetype = "dashed") +
  # points
  geom_point(aes(color=label, shape=library_type)) +
  # labels for points outside ±10%
  geom_text_repel(
    data = root_wide %>% filter(abs(mean_rlen_old - mean_rlen_new) > 0.1 * mean_rlen_new),
    aes(label = library_id), 
    size = 3
  ) +
  labs(
    x = "rlen (new comparison)",
    y = "rlen (old comparison)",
    title = "Comparison of read length between comparisons",
    subtitle = "Shaded band = ±10% difference from identity line"
  )
ggsave("plots/rlen_root.pdf", plot=p)
print(p)
```


Average damage
```{r}
p <- ggplot(root_wide, aes(x = A_new, y = A_old)) +
  # 10% band around y = x
  geom_ribbon(aes(ymin = A_new * 0.9, ymax = A_new * 1.1), 
              fill = "grey80", alpha = 0.5) +
  # y = x line
  geom_abline(color = "blue", linetype = "dashed") +
  # points
  geom_point(aes(color=label, shape=library_type)) +
  # labels for points outside ±10%
  geom_text_repel(
    data = root_wide %>% filter(abs(A_old - A_new) > 0.1 * A_new),
    aes(label = library_id), 
    size = 3
  ) +
  labs(
    x = "A (new comparison)",
    y = "A (old comparison)",
    title = "Comparison of damage (A) between comparisons",
    subtitle = "Shaded band = ±10% difference from identity line"
  )
ggsave("plots/dmg_root.pdf", plot=p)
print(p)
```

# Genus level eukaryotes 

## Number of reads 
Get just genus level, only eukaryotes and name the kingdoms
```{r}
genus <- df %>% filter(rank == "genus") |> # must be classified at genus level 
    filter(grepl("Eukaryota", taxa_path)) |> # must be a Eukaryote 
    mutate(PlantAnimal = case_when( # define plant/animal based on taxpath 
      grepl("Viridiplantae", taxa_path) ~ "plant",
      grepl("Metazoa", taxa_path) ~ "animal",
      grepl("Fungi", taxa_path) ~ "fungi")) %>% 
  filter(!is.na(PlantAnimal))
      
```



Do long to wide with genus df, here we remove taxa path because it is slightly different between the two taxonomy versions
```{r}
genus_wide <- genus %>% select(-taxa_path) %>% 
  pivot_wider(
    names_from = comparison,
    values_from = -c(names(meta), c(comparison,library_id, taxid, name, rank, lca, PlantAnimal))
  )
```


Number of euk reads in old and new 
```{r}
p <- ggplot(genus_wide, aes(x = nreads_new, y = nreads_old, color=label)) +
  geom_abline() + 
  geom_point(alpha=0.7)+ 
  scale_y_continuous(trans="log10") + 
  scale_x_continuous(trans="log10") + 
  facet_wrap(~PlantAnimal, scales = "free")
ggsave("plots/reads_euk.pdf", plot=p)
print(p)
```


More faceted to see clearer

```{r}
p <- ggplot(genus_wide, aes(x = nreads_new, y = nreads_old, color=label)) +
  geom_abline() + 
  geom_point(alpha=0.7)+ 
  scale_y_continuous(trans="log10") + 
  scale_x_continuous(trans="log10") + 
  facet_grid2(vars(PlantAnimal),vars(label),scales="free")
ggsave("plots/reads_euk_facet.pdf", plot=p)
print(p)
```

## Heatmap of differences 
```{r}
read_diff <- genus %>%
  select(name,taxid, library_id, comparison, nreads, PlantAnimal,label) %>%
  pivot_wider(names_from = comparison, values_from = nreads) %>%
  mutate(read_change = new - old)
```

```{r}
for (l in unique(df$label)) {
  
  d <- read_diff %>%
    filter(label == l) %>%
    group_by(PlantAnimal) %>%
    arrange(desc(abs(read_change))) %>%
    slice_head(n = 10)  # top 20 changes

 p <- ggplot(d, aes(x = reorder(name, read_change), y = read_change, fill = PlantAnimal)) +
  geom_col() +
  coord_flip() +
  labs(
    x = "Taxon",
    y = "Read Count Difference",
    title = paste0("Gain/Loss of Reads per Taxon by Library ", l)
  ) 
 
 t = paste0("plots/taxa_gainloss_",l,".pdf")
 ggsave(t, plot=p)
 print(p)
}


```

## Damage plots 

### Big pdf 
We make a giant pdf file called damage_patterns.pdf for the comparisons 
```{r}

plot_pipeline_comparison <- function(df2, only5 = FALSE) {

  # Get info
  nreads_old <- df2 %>% filter(comparison == "old") %>% pull(nreads)
  nreads_new <- df2 %>% filter(comparison == "new") %>% pull(nreads)
  A_b_old <- df2 %>% filter(comparison == "old") %>% pull(A_b)
  A_b_new <- df2 %>% filter(comparison == "new") %>% pull(A_b)
  lib_id <- df2$library_id[1]
  label <- df2$label[1]
  taxon <- df2$name[1]
  taxid <- df2$taxid[1]

  x_cols <- grep("^fwdx\\d+$", names(df2), value = TRUE)
  f_cols <- grep("^fwf\\d+$", names(df2), value = TRUE)

  long_df <- bind_rows(
    df2 %>%
      filter(comparison == "old") %>%
      select(all_of(x_cols), all_of(f_cols)) %>%
      pivot_longer(everything(), names_to = "metric", values_to = "value") %>%
      mutate(position = as.integer(str_extract(metric, "\\d+")),
             type = ifelse(str_starts(metric, "fwdx"), "x", "f"),
             comparison = "old"),
    df2 %>%
      filter(comparison == "new") %>%
      select(all_of(x_cols), all_of(f_cols)) %>%
      pivot_longer(everything(), names_to = "metric", values_to = "value") %>%
      mutate(position = as.integer(str_extract(metric, "\\d+")),
             type = ifelse(str_starts(metric, "fwdx"), "x", "f"),
             comparison = "new")
  ) %>%
    filter(!is.na(value)) %>%
    mutate(type = factor(type, levels = c("f", "x")))

  if (only5) {
    long_df <- long_df %>% filter(position < 7)
  }

  A_points <- tibble(
    position = 0,
    value = c(A_b_old, A_b_new),
    type = "A",
    comparison = c("old", "new")
  )

  title <- paste0(
    taxon, " (taxid: ", taxid, ")\n",
    "Library: ", lib_id, ", Label: ", label, "\n",
    "nreads (old/new): ", nreads_old, " / ", nreads_new,
    ", A_b: ", round(A_b_old, 3), " / ", round(A_b_new, 3)
  )

  ggplot(long_df, aes(x = position, y = value, color = comparison, linetype = type)) +
    geom_line() +
    geom_point(data = filter(long_df, type == "f")) +
    geom_point(data = A_points, aes(shape = type), size = 3, stroke = 1.2, inherit.aes = TRUE) +
    scale_color_manual(values = c("old" = "royalblue4", "new" = "tomato4")) +
    scale_linetype_manual(values = c("x" = "solid", "f" = "dashed")) +
    scale_shape_manual(values = c(A = 8)) +
    labs(
      title = title,
      x = "Position from 5' end",
      y = "Damage Estimate",
      color = "comparison",
      linetype = "Metric",
      shape = NULL
    ) 
}

```

```{r}
# Step 1: Find taxon-library combinations with >= 100 reads in BOTH comparisons
valid_taxa <- genus %>%
  group_by(name, taxid, library_id) %>%
  filter(all(c("old", "new") %in% comparison)) %>%
  summarise(
    n_old = nreads[comparison == "old"],
    n_new = nreads[comparison == "new"],
    .groups = "drop"
  ) %>%
  filter(n_old >= 100, n_new >= 100)

# Step 2: Filter big df
genus_filtered <- semi_join(genus, valid_taxa, by = c("name", "taxid", "library_id"))

# Step 3: Split into groups
groups <- genus_filtered %>%
  group_by(name, taxid, library_id) %>%
  group_split()

# Step 4: Create and save all plots
plots <- list()

for (df2 in groups) {
  p <- plot_pipeline_comparison(df2)
    plots <- append(plots, list(p))
}

# Save only if we have valid plots
if (length(plots) > 0) {
  pdf("damage_patterns.pdf", width = 8, height = 6)
  for (p in plots) {
    print(p)
  }
  dev.off()
} else {
  message("No valid plots generated — PDF not created.")
}

```


### Just damage estimates 

```{r}
p <- genus_wide %>% filter(nreads_new >= 100 & nreads_old >= 100) %>%
ggplot(aes(x=A_new, y = A_old, color = label)) +
  geom_abline() + 
  geom_point(alpha = 0.7) +
  facet_wrap(~PlantAnimal)
ggsave("plots/dmg_euk.pdf", plot=p)
print(p)  
```

## Number of reads total for plant and animal
```{r}

id_df <- genus %>% group_by(library_id, label, PlantAnimal,comparison) %>%
  summarise(total_nreads = sum(nreads)) %>% 
  pivot_wider(names_from = comparison, values_from = total_nreads)


p <- ggplot(id_df, aes(x=new, y =old, color=label)) +
  geom_abline() + 
  geom_point() + 
  facet_wrap(~PlantAnimal, scales="free")
ggsave("plots/reads_total_euk.pdf", plot=p)
print(p)  
```


# Number of taxa identified
```{r}

id_df <- genus %>% group_by(library_id, label, PlantAnimal,comparison) %>%
  summarise(num = n()) %>% 
  pivot_wider(names_from = comparison, values_from = num)


p <- ggplot(id_df, aes(x=new, y =old, color=label)) +
  geom_abline() + 
  geom_point() + 
  facet_wrap(~PlantAnimal, scales="free")
ggsave("plots/taxa_euk.pdf", plot=p)
print(p)  
```
# Taxa

Which taxa are unique? Do Venn diagram 
```{r}


library_ids <- unique(genus$library_id)
plots <- list()

for (lib in library_ids) {
  df_lib <- genus %>%
    filter(library_id == lib)

  label <- unique(df_lib$label)

  # Loop over PlantAnimal categories
  venn_plots <- list()
  for (group in unique(df_lib$PlantAnimal)) {
    df_group <- df_lib %>%
      filter(PlantAnimal == group)

    taxa_old <- df_group %>%
      filter(comparison == "old") %>%
      pull(name) %>%
      unique()

    taxa_new <- df_group %>%
      filter(comparison == "new") %>%
      pull(name) %>%
      unique()

    venn_counts <- list(
      old = taxa_old,
      new = taxa_new
    )

    fit <- tryCatch(euler(venn_counts), error = function(e) NULL)

    if (!is.null(fit)) {
      p <- plot(
        fit,
        fills = list(fill = c("royalblue4", "tomato4"), alpha = 0.5),
        quantities = TRUE,
        main = paste("Library:", lib, "|", group)
      )
      venn_plots <- append(venn_plots, list(p))
    }
  }

  if (length(venn_plots) > 0) {
    combined <- do.call(gridExtra::arrangeGrob, c(venn_plots, ncol = 2,
      top = paste("Taxa overlap for library:", lib, "-", label)))
    plots <- append(plots, list(combined))
  }
}

# Save to PDF
if (length(plots) > 0) {
  pdf("venn_taxa_by_library.pdf", width = 10, height = 8)
  for (p in plots) {
    grid::grid.draw(p)
    grid::grid.newpage()
  }
  dev.off()
} else {
  message("No valid plots generated — PDF not created.")
}


```

# PCoA

Do PCoA, three plots per label (plants, animals, fungi), each dot is one library
```{r}


plots <- list()

for (lab in unique(genus$label)) {

  for (group in c("plant", "animal", "fungi")) {

    genus_sub <- genus %>%
      filter(label == lab, PlantAnimal == group, !is.na(name), !is.na(nreads)) %>%
      mutate(
        taxa = paste0(name, "-", taxid),
        sample = paste(library_id, comparison, sep = "_")
      )

    # Skip if too few samples
    if (length(unique(genus_sub$sample)) < 3) next

    # Pivot to wide
    tmp_dt <- as.data.table(genus_sub)
    wide_counts <- dcast(tmp_dt, taxa ~ sample, value.var = "nreads", fill = 0)

    if (nrow(wide_counts) < 2) next

    rownames(wide_counts) <- wide_counts$taxa
    wide_data <- as.data.frame(wide_counts[, -1])

    # Compute distances and PCoA
    dist_matrix <- vegdist(t(wide_data), method = "bray")
    pcoa_result <- cmdscale(dist_matrix, k = 2, eig = TRUE)

    pcoa_df <- data.frame(pcoa_result$points)
    colnames(pcoa_df) <- c("PCoA1", "PCoA2")
    pcoa_df$sample <- rownames(pcoa_result$points)

    # Join metadata
    pcoa_df <- pcoa_df %>%
      separate(sample, into = c("library_id", "comparison"), sep = "_", remove = FALSE) %>%
      left_join(
        genus_sub %>% 
          select(library_id, comparison) %>%
          distinct(),
        by = c("library_id", "comparison")
      )

    # Plot
    p <- ggplot(pcoa_df, aes(x = PCoA1, y = PCoA2, color = comparison)) +
      geom_path(aes(group = library_id), color = "grey50", linewidth = 0.5, alpha = 0.7) +
      geom_point(size = 4) +
      labs(
        title = paste("PCoA | Label:", lab, "| Group:", group),
        x = "PCoA1",
        y = "PCoA2",
        color = "Pipeline"
      ) +
      theme_test(base_size = 15)

    plots <- append(plots, list(p))
  }
}

# Save all to PDF
if (length(plots) > 0) {
  pdf("pcoa.pdf", width = 10, height = 8)
  for (p in plots) print(p)
  dev.off()
} else {
  message("No valid PCoA plots generated.")
}

```

```{r}
```

```{r}
```

```{r}
```

```{r}
```